### 2010-02-09 paper/abstract.tex
Abstract was accepted.  Now we gotta deliver.  I got this.

\documentclass[12pt]{article}
\oddsidemargin 0em
\evensidemargin 0em
\topmargin 0em
\textwidth 40em
\parindent 0pt

\begin{document}

\subsubsection*{Drawing Trees and Animating Tree Changes}

Sandro Badame, Peter Boothe$^*$, Manhattan College
\medskip

Drawing trees and animating tree changes (subtree addition, replacement, and
removal) can allow new visualizations of computational and graph theoretic
processes.  We will be demonstrating a library that we have designed and built
for exactly this purpose as well as two visualizations designed to show off its
capabilities.  Our first demonstration shows the ability to watch the execution
of a piece of Lisp code, while our second example shows a tree being parsed by
a tree automaton.  In the Lisp example, the code being executed may be entered
by the user, and in the tree example the tree is user-entered, although the
automaton is not.  Drawing a graph as a tree can even prove useful when the
graph being drawn is neither directed nor acyclic.  We present one last
example visualization using a graph which is not a DAG, which allows us to
animate the creation and usage of a doubly-linked list.  We hope that the
library will be of use to educators who want to provide a visualization of
trees and graphs and of computational processes on these structures.  
\medskip

Keywords:  computing, teaching, graph visualization, trees, animation
\end{document}

### 2010-03-05 src/graphics/TreeDisplay.java
Binary tree insertion now works.

@@ -101,6 +101,8 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
             }
 
             t = curr_layer.remove();
+            if (t == null) continue;
+
             t.addTreeChangeListener(this);
             Node n = new Node(t, xpos, (layer+1)*50, cachedGraphics);
             xpos += n.width + PADDING;

### 2010-03-05 src/main/BinaryTreeInsertion.java

@@ -36,7 +36,7 @@ public class BinaryTreeInsertion extends LispFreeWindow{
   
   private void insertNumber(int num, Tree t){
         
-    if (t.getData().isEmpty()){
+    if (t.getData() == null || t.getData().isEmpty()){
       t.setData(num, false);
       t.addChild(new Tree("", null), false);
       t.addChild(new Tree("", null));
@@ -69,54 +69,109 @@ public class BinaryTreeInsertion extends LispFreeWindow{
 
   private void rightRotation(Tree root){
     Tree c = root;
+    String cdata = c.getData();
     Tree b = c.getChild(LEFT);
     Tree a = b.getChild(LEFT);
-    
-    Tree left = a;
-    Tree right = new Tree(c.getData(), Arrays.asList(new Tree[]{b.getChild(RIGHT), c.getChild(RIGHT)}));
-    root.setData(b.getData());
-    root.removeChildren();
-    root.addChild(left);
-    root.addChild(right);
+    Tree T0 = a.getChild(LEFT);
+    Tree T1 = a.getChild(RIGHT);
+    Tree T2 = b.getChild(RIGHT);
+    Tree T3 = c.getChild(RIGHT);
+
+    root.setData(b.getData(), false);
+    b.setData(cdata, false);
+    c = b;
+    b = root;
+
+    a.removeChildren(false);
+    b.removeChildren(false);
+    c.removeChildren(false);
+    a.addChild(T0, false);
+    a.addChild(T1, false);
+    b.addChild(a, false);
+    b.addChild(c, false);
+    c.addChild(T2, false);
+    c.addChild(T3);
   }
   
   private void leftRotation(Tree root) {
     Tree a = root;
     Tree b = a.getChild(RIGHT);
     Tree c = b.getChild(RIGHT);
+
+    Tree T0 = a.getChild(LEFT);
+    Tree T1 = b.getChild(LEFT);
+    Tree T2 = c.getChild(LEFT);
+    Tree T3 = c.getChild(RIGHT);
     
-    Tree left = new Tree(a.getData(), Arrays.asList(new Tree[]{a.getChild(LEFT), b.getChild(LEFT)}));
-    Tree right = c;
-    root.setData(b.getData());
-    root.removeChildren();
-    root.addChild(left);
-    root.addChild(right);
+    String adata = a.getData();
+    root.setData(b.getData(), false);
+    a = b;
+    b = root;
+
[…]

### 2010-03-05 src/main/Tree.java

@@ -4,10 +4,11 @@ import java.util.*;
 import jscheme.*;
 
 public class Tree {
+    protected Tree parent = null;
     private String data = null;
     private Vector<Tree> children = new Vector<Tree>();
     
-    Vector<TreeChangeListener> listeners = new Vector<TreeChangeListener>();
+    Set<TreeChangeListener> listeners = new HashSet<TreeChangeListener>();
 
     public Tree() {}
     public Tree(Object data, Iterable<Tree> kids) 
@@ -109,7 +110,14 @@ public class Tree {
     public void addChild(Tree child, boolean fire)
     {
         synchronized (children) {
+            child.parent = this;
             children.add(child);
+            depthcache = -1;
+            Tree t = parent;
+            while (t != null && t.depthcache != -1) {
+                t.depthcache = -1;
+                t = t.parent;
+            }
         }
 
         synchronized (listeners) {
@@ -122,13 +130,19 @@ public class Tree {
 
     public void removeChildren(){ removeChildren(true); }
     public void removeChildren(boolean fire){
-        synchronized (children) { children.clear(); }
+        synchronized (children) { 
+            for (Tree t : children) t.parent = null;
+            children.clear(); 
+        }
+
         if (fire) fireChildrenClearedEvent();
     }
 
+    int depthcache = -1;
     public int depth()
     {
         synchronized(children){
+          if (depthcache != -1) return depthcache;
           int depth = 1;
           if (children != null) {
               for (Tree kid : children) {
@@ -141,7 +155,7 @@ public class Tree {
                   }
               }
           }
-          return depth;
+          return depthcache = depth;
         }
     }
 

### 2010-03-08 src/main/DemoMenu.java
Added a git ignore, added a demo menu, run target now runs the demo menu, the lisp example target has been renamed to runlisp

@@ -1,0 +1,27 @@
+package main;
+
+import javax.swing.JOptionPane;
+
+public class DemoMenu{
+	public static void main(String[] args){
+		Object[] options = new String[]{"Scheme", "AVL Trees", "Tree Automata"};
+		String s = (String)JOptionPane.showInputDialog(
+                    null, "The following demos are available",
+		    "Demo Menu", JOptionPane.PLAIN_MESSAGE, null,
+                    options, "AVL Trees");
+
+		//If a string was returned, say so.
+		if ((s != null) && (s.length() > 0)) {
+			if (s.equals("Scheme")){
+				MainWindow.main(args);
+			}else if (s.equals("AVL Trees")){
+				BinaryTreeInsertion.main(args);
+			}else if (s.equals("Tree Automata")){
+				TreeAutomaton.main(args);
+			}else{
+				System.out.println("DemoMenu is broken");
+			}	
+		    return;
+		}
+	}
+}

### 2010-03-08 presentation/talk.tex
Cleanup and fixup.  Also, the talk is in.

\documentclass{beamer}
\usetheme{manhattan}  % Now it's a beamer presentation with the UO theme!

\usepackage{url}

% Make a new command that will make a new subsection and a frame with the same title
\newcommand{\fst}[2]{\subsection{#1}\frame{\frametitle{#1} #2}}

\title{Drawing Trees and Animating Tree Changes}
\author[Badame and Boothe]{Sandro Badame \and Peter Boothe*}
\date{9 March 2010}
\institute[Manhattan College]{
    \url{{sbadame.student,peter.boothe}@manhattan.edu}\\
    Mathematics and Computer Science \\
    Manhattan College}

\begin{document}
\frame{\titlepage}

\section{Introduction}

\fst{The Problem}{
\begin{itemize}
\item Trees undergoing changes occur in many contexts
    \begin{itemize}
        \item Visualization can be a help 
    \end{itemize}
\item Writing graphics and layout code from the ground up is hard.
\item Writing animation code is harder
\item Worst of all often the people who have skills with algorithms have little 
  experience with graphics programming.
    \begin{itemize}
        \item And visa-versa
    \end{itemize}
\end{itemize}
}

\fst{Our Solution}{
\begin{itemize}
\item We have built a Java library for displaying trees and animating tree changes
\item To demonstrate the generality of the library we have created three example visualizations.
    \begin{enumerate}
        \item AVL Trees
        \item Automata
        \item Scheme
    \end{enumerate}
\end{itemize}
}

\section{Demonstrations}

\fst{AVL Trees}{
\begin{itemize}
    \item Example of self-adjusting binary search trees
    \item AVL Tree invariant: The left and right subtree of a node differ in maximum depth by no more than 1
    \item If this invariant is violated, fix the tree via rotations
\end{itemize}
}

\fst{Tree Automata}{
\begin{itemize}
    \item Take as input trees rather than strings.
    \item Progressively rewrite the input tree until either
    \begin{itemize}
        \item There is no rule to transform the tree (reject) \\
            ~~~~OR
        \item The tree is isomorphic to a member of the set of base trees  (accept)
    \end{itemize}
\end{itemize}
}

\frame{
\frametitle{Our Tree Automaton: Tautology Checker}
    \begin{description}
        \item{\bf Base Trees:} $$\{ q1 \}$$

        \item{\bf Rewrite Rules:}

~\\

    \hspace{-.8in}
      \begin{tabular}{c | c | c}
         $1 \to q1$         & $(AND ~q1 ~q0) \to q0$ & $(OR ~q1 ~q0) \to q1$ \\
         $0 \to q0$         & $(AND ~q1 ~q1) \to q1$ & $(OR ~q1 ~q1) \to q1$ \\
         $(NOT ~q1) \to q0$ & $(AND ~q0 ~q0) \to q0$ & $(OR ~q0 ~q0) \to q0$ \\
         $(NOT ~q0) \to q1$ & $(AND ~q0 ~q1) \to q0$ & $(OR ~q0 ~q1) \to q1$ 
        \end{tabular}
    \end{description}
}

\fst{Scheme/Lisp}{
    \begin{itemize} 
        \item Scheme/Lisp code represents the Abstract Syntax Tree directly.
        \item Executing Scheme code corresponds to manipulation of the AST

           $\therefore$ We can animate the execution of Scheme code
    \end{itemize}
}

\frame{
    \frametitle{Summary of Examples}
    \begin{itemize} 
        \item These animations where not too hard to write.
        \item The extra code required to perform the animation was on the order of 10 lines of Java code. 
        \item The same library is used for all animations.
        \item We feel that we have created a library that can be useful in a wide variety of situations.
    \end{itemize} 
}

\section{Wrap-Up}

\fst{Conclusion}{
    \begin{itemize} 
        \item Write visualizations for yourself 
        \item Get your students to write them
        \item \url{http://github.com/pboothe/Lispy-Animator}
        \item Any Questions?
    \end{itemize} 
}

\end{document}

### 2010-03-08 src/main/BinaryTreeInsertion.java

@@ -36,6 +36,7 @@ public class BinaryTreeInsertion extends LispFreeWindow{
   
   private void insertNumber(int num, Tree t){
         
+    System.out.println("adding num: " + num + " to " + t);
     if (t.getData() == null || t.getData().isEmpty()){
       t.setData(num, false);
       t.addChild(new Tree("", null), false);
@@ -132,6 +133,7 @@ public class BinaryTreeInsertion extends LispFreeWindow{
     
     String cdata = c.getData();
     root.setData(b.getData(), false);
+    b.setData(cdata, false);
     c = b;
     b = root;
     
@@ -157,9 +159,10 @@ public class BinaryTreeInsertion extends LispFreeWindow{
     Tree T2 = b.getChild(RIGHT);
     Tree T3 = c.getChild(RIGHT);
     
-    String cdata = c.getData();
+    String adata = a.getData();
     root.setData(b.getData(), false);
-    c = b;
+    b.setData(adata, false);
+    a = b;
     b = root;
     
     a.removeChildren(false);

### 2010-03-08 src/main/Tree.java

@@ -160,14 +160,19 @@ public class Tree {
     }
 
     public void addTreeChangeListener(TreeChangeListener listener){
+        addTreeChangeListener(listener, new HashSet<Tree>());
+    }
+
+    public void addTreeChangeListener(TreeChangeListener listener, Set<Tree> heard){
         synchronized (listeners) {
             listeners.add(listener);
         }
 
         synchronized (children) {
             for (Tree kid : children) {
-                if (kid == null) continue;
-                kid.addTreeChangeListener(listener);
+                if (kid == null || heard.contains(kid)) continue;
+                heard.add(kid);
+                kid.addTreeChangeListener(listener, heard);
             }
         }
     }

### 2010-03-08 presentation/talk.tex
Modifications to the talk.

@@
 \documentclass{beamer}
-\usetheme{manhattan}  % Now it's a beamer presentation with the UO theme!
+\usetheme{manhattan}  % Now it's a beamer presentation with the Manhattan College theme!
 
@@
     \begin{itemize}
-        \item And visa-versa
+        \item And vice-versa
     \end{itemize}
@@
     \end{enumerate}
+\end{itemize}
+}
+
+\fst{How It Works}{
+    \begin{itemize} 
+        \item Each vertex is a magnet that repulses its neighbors
+        \item Each edge is a spring, pulling adjacent vertices closer
+        \item Each vertex has a fixed vertical position and a fixed horizontal order
     \end{itemize} 
@@
 
-\section{Wrap-Up}
+\section{Wrapping Up}
 

### 2010-03-08 src/main/Lisp.java
LISP works.

@@ -4,7 +4,9 @@ import java.util.*;
 
 
 public class Lisp {
-  String reservedwords[] = { "+", "-", "*", "/", "<", ">", "if", "fun" };
+  private static final String FUN = "fun";
+  private static final String IF = "if";
+  String reservedwords[] = { "+", "-", "*", "/", "<", ">", "=", IF, FUN};
 
   Tree root = new Tree();
   public Lisp(String code) throws CompilationException
@@ -107,19 +109,19 @@ public class Lisp {
 
         kids.remove(0);
         String word = first.getData();
-        if (word.equals("fun")) {
-            // (fun name arg1 arg2 [...] body) ; note that args are optional
+        if (word.equals(FUN)) {
+            // (FUN (arg1 arg2 [...]) body) ; note that args are optional
             if (kids.size() <= 1) 
                 throw new CompilationException(code);
             if (kids.get(1).getChildren() == null)
                 throw new CompilationException(code);
 
-            return new Tree("fun", kids);
-        } else if (word.equals("if")) {
+            return new Tree(FUN, kids);
+        } else if (word.equals(IF)) {
             // (if cond then else)
             if (kids.size() != 3) 
                 throw new CompilationException(code);
-            return new Tree("if", kids);
+            return new Tree(IF, kids);
         } else {
             return new Tree(word, kids);
         }
@@ -128,6 +130,182 @@ public class Lisp {
     }
   }
 
+  private static void replace(Tree code, String data, Tree replacement)
+  {
+      System.out.println("Trying to replace" + data + " with " + replacement + " in " + code);
+      if (code.getData().equals(data)) {
+          System.out.println("Replacing " + data + " with " + replacement);
+          code.removeChildren(false);
+          for (Tree t : replacement.getChildren()) {
+              code.addChild(new Tree(t), false);
+          }
+          code.setData(replacement.getData());
+      } else if (code.getData().equals(FUN)) {
+          //Shadowing
+          Vector<Tree> fn = code.getChildren();
+          Vector<Tree> args = fn.get(0).getChildren();
+
+          for (Tree t : args) {
+              if (t.getData().equals(data)) {
+                  return;
+              }
+          }
+
+          // If we made it here, the replacement term is not in the args
+          replace(fn.get(1), data, replacement);
+      } else {
+          for (Tree t : code.getChildren()) {
[…]

### 2010-03-08 src/main/MainWindow.java

@@ -7,8 +7,6 @@ import java.io.*;
 import java.util.*;
 import javax.swing.*;
 
-import jscheme.*;
-
 public class MainWindow extends JFrame {
 
   private TreeDisplay treeDisplay = new TreeDisplay();
@@ -34,6 +32,20 @@ public class MainWindow extends JFrame {
       public void actionPerformed(ActionEvent e){ step(); }
     });
 
+    bottomInput.setPlayAllButtonAction(new ActionListener(){
+      public void actionPerformed(ActionEvent e){ 
+        new Thread() {
+            public void run() {
+                while (step()) {
+                  try {
+                    Thread.sleep(1000);
+                  } catch (InterruptedException ie) {}
+                }
+            }  
+        }.start();
+      }
+    });
+
     bottomInput.setChangeAction(new ActionListener(){
       public void actionPerformed(ActionEvent e){ 
           outputtext.setText(bottomInput.getText().replace('\n', ' '));
@@ -44,53 +56,29 @@ public class MainWindow extends JFrame {
     pack();
   }
 
+  Lisp lisp = null;
   synchronized private void startLayout()
   {
-      String code = outputtext.getText();
-      int i = code.lastIndexOf("\n");
-      code = code.substring(i+1);
-      Object lisp =  new InputPort(new StringReader(code)).read();
-      
-      if (lisp == InputPort.EOF) {
+      String code = bottomInput.getText().trim();
+      try {
+          lisp = new Lisp(code);
+          treeDisplay.setTree(lisp.root);
+          System.out.println(lisp.toString());
+      } catch (CompilationException ce) {
         treeDisplay.setMessage("Problem with: " + code);
         System.out.println("Problem with: " + code);
-      }else{
-        treeDisplay.setTree(fromLisp(lisp));
-        System.out.println(lisp.toString());
       }
   }
 
-  private static Tree fromLisp(Object lisp)
-  {
-      if (lisp instanceof Pair) {
-          Pair p = (Pair)lisp;
-
-          Tree t = new Tree("", Arrays.asList(new Tree[] { fromLisp(p.first) }));
-          while (p.rest instanceof Pair) {
-              p = (Pair)p.rest;
-              t.addChild(fromLisp(p.first));
-          }
-
-          return t;
-      } else 
-          return new Tree(lisp);
-  }
-
-  private void step()
+  private boolean step()
   {
[…]

### 2010-03-08 src/main/Tree.java

@@ -21,6 +21,14 @@ public class Tree {
             }
     }
     public Tree(Iterable<Tree> kids) { this(null, kids); }
+    public Tree(Tree copy)
+    {
+        setData(copy.getData(), false);
+        for (Tree t : copy.getChildren()) {
+            addChild(new Tree(t), false);
+        }
+    }
+
     public Tree(Object o) {
         this(o, null);
     }

### 2010-03-08 src/main/Tree.java
Factorial targe established and Tree now updates depth on child removal

@@ -141,6 +141,13 @@ public class Tree {
         synchronized (children) { 
             for (Tree t : children) t.parent = null;
             children.clear(); 
+
+            depthcache = -1;
+            Tree t = parent;
+            while (t != null && t.depthcache != -1) {
+                t.depthcache = -1;
+                t = t.parent;
+            }
         }
 
         if (fire) fireChildrenClearedEvent();

### 2010-03-10 src/graphics/TreeDisplay.java
This is pretty much what we presented.

@@ -121,7 +121,6 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
     @Override
     public void paint(Graphics g)
     {
-        synchronized (this) {
             Graphics2D g2 = graphicsPrep(g);
             if (message != null){
                 g.drawString(message, 0, getHeight()/2);
@@ -129,6 +128,7 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
             } 
 
             if (treechanged) {
+                System.out.println("TREE CHANGED");
                 treechanged = false;
                 layout(tree, 0, PADDING);
             }
@@ -148,7 +148,6 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
 
 
             g2.dispose();
-        }
     }
 
     public double adjust()
@@ -254,6 +253,7 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
     	animations.offer(new Animation() {
                     void animate()
                     {
+                        System.out.println("Animating: " + superTree);
                         // Make the new layout
                         TreeDisplay td = new TreeDisplay(false);
                         td.cachedGraphics = cachedGraphics;
@@ -328,6 +328,7 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
                         }
                         
                         repaint();
+                        System.out.println("Done animating: " + superTree);
                     }
                 });
     }

### 2010-03-10 src/main/BinaryTreeInsertion.java

@@ -1,5 +1,7 @@
 package main;
 
+import java.awt.event.ActionEvent;
+import java.awt.event.ActionListener;
 import java.util.Arrays;
 import java.util.Vector;
 
@@ -13,6 +15,12 @@ public class BinaryTreeInsertion extends LispFreeWindow{
   
   public BinaryTreeInsertion(){
     super("Binary Tree Insertion");
+
+    bottomInput.setPlayAllButtonAction(new ActionListener() {
+            public void actionPerformed(ActionEvent e){ 
+                treeDisplay.setTree(tree = getStartingTree());
+            }
+        });
   }
   
   protected Tree getStartingTree(){
@@ -23,14 +31,19 @@ public class BinaryTreeInsertion extends LispFreeWindow{
     if (bottomInput.getText().isEmpty() == false){
       System.out.println(bottomInput.getText());
       String[] numbers = bottomInput.getText().split(",");
-      for(String num : numbers){
-        try{
-          insertNumber(Integer.parseInt(num.trim()), tree);
-        }catch(NumberFormatException nfe){
+      try{
+          for(String num : numbers){
+              insertNumber(Integer.parseInt(num.trim()), tree);
+                
+              //try { Thread.sleep(2000); } catch (InterruptedException ie) {}
+          }
+
+          tree.setData(tree.getData());
+
+          bottomInput.text.setText("");
+      }catch(NumberFormatException nfe){
           JOptionPane.showMessageDialog(this, "Please enter a list of comma seperated numbers");
-        }
       }
-      bottomInput.text.setText("");
     }
   }
   
@@ -43,10 +56,14 @@ public class BinaryTreeInsertion extends LispFreeWindow{
       t.addChild(new Tree("", null));
     }else if ( num  < Integer.parseInt(t.getData())){
       insertNumber(num, t.getChild(LEFT));
+      System.out.println("About to rebalance " + t);
       rebalance(t);
+      System.out.println("done rebalancing " + t);
     }else{
       insertNumber(num, t.getChild(RIGHT));
+      System.out.println("About to rebalance " + t);
       rebalance(t);
+      System.out.println("done rebalancing " + t);
     }
   }
   
@@ -70,7 +87,6 @@ public class BinaryTreeInsertion extends LispFreeWindow{
 
[…]

### 2010-03-10 src/main/MainWindow.java

@@ -38,7 +38,7 @@ public class MainWindow extends JFrame {
             public void run() {
                 while (step()) {
                   try {
-                    Thread.sleep(1000);
+                    Thread.sleep(2000);
                   } catch (InterruptedException ie) {}
                 }
             }  

### 2010-03-10 src/jscheme/Environment.java
No more jscheme...

@@ -1,100 +1,0 @@
-package jscheme;
-
-/** Environments allow you to look up the value of a variable, given
- * its name.  Keep a list of variables and values, and a pointer to
- * the parent environment.  If a variable list ends in a symbol rather
- * than null, it means that symbol is bound to the remainder of the
- * values list. 
- * @author Peter Norvig, peter@norvig.com http://www.norvig.com 
- * Copyright 1998 Peter Norvig, see http://www.norvig.com/license.html */
-
-public class Environment extends SchemeUtils {
-    public Object vars;
-    public Object vals;
-    public Environment parent;
-    
-    /** A constructor to extend an environment with var/val pairs. */
-    public Environment(Object vars, Object vals, Environment parent) {
-        this.vars = vars;
-        this.vals = vals;
-        this.parent = parent;
-	if (!numberArgsOK(vars, vals))
-	    warn("wrong number of arguments: expected " + vars +
-			" got " + vals);
-    }
-
-    /** Construct an empty environment: no bindings. **/
-    public Environment() {}
-
-    /** Find the value of a symbol, in this environment or a parent. */
-    public Object lookup (String symbol) {
-	Object varList = vars, valList = vals;
-	// See if the symbol is bound locally
-	while (varList != null) {
-	    if (first(varList) == symbol) {
-		return first(valList);
-	    } else if (varList == symbol) {
-		return valList;
-	    } else {
-		varList = rest(varList);
-		valList = rest(valList);
-	    }
-	}
-	// If not, try to look for the parent
-	if (parent != null) return parent.lookup(symbol);
-	else return error("Unbound variable: " + symbol);
-    }
-    
-    /** Add a new variable,value pair to this environment. */
-     public Object define(Object var, Object val) {
-	 vars = cons(var, vars);
-	 vals = cons(val, vals);
-	 if (val instanceof Procedure 
-	     && ((Procedure)val).name.equals("anonymous procedure"))
-	     ((Procedure)val).name = var.toString();
-	 return var;
-     }
-
-  /** Set the value of an existing variable **/
-  public Object set(Object var, Object val) {
-    if (!(var instanceof String)) 
-      return error("Attempt to set a non-symbol: " 
-			  + stringify(var));;
-    String symbol = (String) var;
-    Object varList = vars, valList = vals;
-    // See if the symbol is bound locally
-    while (varList != null) {
-      if (first(varList) == symbol) {
[…]

### 2010-03-10 src/jscheme/InputPort.java

@@ -1,210 +1,0 @@
-package jscheme;
-import java.io.*;
-
-/** InputPort is to Scheme as InputStream is to Java. 
- * @author Peter Norvig, peter@norvig.com http://www.norvig.com  
- * Copyright 1998 Peter Norvig, see http://www.norvig.com/license.html **/
-
-public class InputPort extends SchemeUtils {
-
-  public static  String EOF = "#!EOF";
-  boolean isPushedToken = false;
-  boolean isPushedChar = false;
-  Object  pushedToken = null;
-  int     pushedChar = -1;
-  Reader  in;
-  StringBuffer buff = new StringBuffer();
-
-  /** Construct an InputPort from an InputStream. **/
-  public InputPort(InputStream in) { this.in = new InputStreamReader(in);}
-
-  /** Construct an InputPort from a Reader. **/
-  public InputPort(Reader in) { this.in = in;}
-
-  /** Read and return a Scheme character or EOF. **/
-  public Object readChar() {
-    try {
-      if (isPushedChar) {
-	isPushedChar = false;
-	if (pushedChar == -1) return EOF; else return chr((char)pushedChar);
-      } else {
-	int ch = in.read();
-	if (ch == -1) return EOF; else return chr((char)ch);
-      }
-    } catch (IOException e) {
-      warn("On input, exception: " + e);
-      return EOF;
-    }
-  }
-
-  /** Peek at and return the next Scheme character (or EOF).
-   * However, don't consume the character. **/
-  public Object peekChar() {
-    int p = peekCh();
-    if (p == -1) return EOF; else return chr((char)p);
-  }
-
-  /** Push a character back to be re-used later. **/
-  int pushChar(int ch) {
-    isPushedChar = true;
-    return pushedChar = ch;
-  }
-
-  /** Pop off the previously pushed character. **/
-  int popChar() {
-    isPushedChar = false;
-    return pushedChar;
-  }
-
-  /** Peek at and return the next Scheme character as an int, -1 for EOF.
-   * However, don't consume the character. **/
-  public int peekCh() {
-    try { return isPushedChar ? pushedChar : pushChar(in.read()); }
-    catch (IOException e) {
-      warn("On input, exception: " + e);
-      return -1;
-    }
-  }
-
-  /** Read and return a Scheme expression, or EOF. **/
-  public Object read() {
-    try {
-      Object token = nextToken(); 
-      if (token == "(")
-	return readTail(false);
-      else if (token == ")")
[…]

### 2010-03-10 src/jscheme/JavaMethod.java

@@ -1,88 +1,0 @@
-package jscheme;
-import java.lang.reflect.*;
-
-/** @author Peter Norvig, peter@norvig.com http://www.norvig.com
- * Copyright 1998 Peter Norvig, see http://www.norvig.com/license.html */
-
-public class JavaMethod extends Procedure {
-
-  Class[] argClasses;
-  Method method;
-  boolean isStatic;
-
-  public JavaMethod(String methodName, Object targetClassName, 
-		    Object argClassNames) {
-    this.name = targetClassName + "." + methodName;
-    try {
-      argClasses = classArray(argClassNames);
-      method = toClass(targetClassName).getMethod(methodName, argClasses);
-      isStatic = Modifier.isStatic(method.getModifiers());
-    } catch (ClassNotFoundException e) { 
-      error("Bad class, can't get method " + name); 
-    } catch (NoSuchMethodException e) { 
-      error("Can't get method " + name); 
-    }
-    
-  }
-
-  /** Apply the method to a list of arguments. **/
-  public Object apply(Scheme interpreter, Object args) {
-    try {
-      if (isStatic) return method.invoke(null, toArray(args));
-      else return method.invoke(first(args), toArray(rest(args)));
-    } catch (IllegalAccessException e) { ; }
-    catch (IllegalArgumentException e) { ; }
-    catch (InvocationTargetException e) { ; }
-    catch (NullPointerException e) { ; }
-    return error("Bad Java Method application:" + this 
-			+ stringify(args) + ", "); 
-  }
-
-  public static Class toClass(Object arg) throws ClassNotFoundException { 
-    if      (arg instanceof Class)  return (Class)arg;
-    arg = stringify(arg, false);
-
-    if (arg.equals("void"))    return java.lang.Void.TYPE;
-    else if (arg.equals("boolean")) return java.lang.Boolean.TYPE;
-    else if (arg.equals("char"))    return java.lang.Character.TYPE;
-    else if (arg.equals("byte"))    return java.lang.Byte.TYPE;
-    else if (arg.equals("short"))   return java.lang.Short.TYPE;
-    else if (arg.equals("int"))     return java.lang.Integer.TYPE;
-    else if (arg.equals("long"))    return java.lang.Long.TYPE;
-    else if (arg.equals("float"))   return java.lang.Float.TYPE;
-    else if (arg.equals("double"))  return java.lang.Double.TYPE;
-    else return Class.forName((String)arg);
-  }
-
[…]

### 2010-03-10 src/jscheme/Macro.java

@@ -1,33 +1,0 @@
-package jscheme;
-
-/** @author Peter Norvig, peter@norvig.com http://www.norvig.com 
- * Copyright 1998 Peter Norvig, see http://www.norvig.com/license.html **/
-
-public class Macro extends Closure {
-
-    /** Make a macro from a parameter list, body, and environment. **/
-    public Macro (Object parms, Object body, Environment env) {
-      super(parms, body, env);
-    }
-
-  /** Replace the old cons cell with the macro expansion, and return it. **/
-  public Pair expand(Scheme interpreter, Pair oldPair, Object args) {
-    Object expansion = apply(interpreter, args);
-    if (expansion instanceof Pair) {
-      oldPair.first = ((Pair)expansion).first;
-      oldPair.rest  = ((Pair)expansion).rest;
-    } else {
-      oldPair.first = "begin";
-      oldPair.rest = cons(expansion, null);
-    }
-    return oldPair;
-  }
-
-  /** Macro expand an expression **/
-  public static Object macroExpand(Scheme interpreter, Object x) {
-    if (!(x instanceof Pair)) return x;
-    Object fn = interpreter.eval(first(x), interpreter.globalEnvironment);
-    if (!(fn instanceof Macro)) return x;
-    return ((Macro)fn).expand(interpreter, (Pair)x, rest(x));
-  }
-}

### 2010-03-10 src/jscheme/Pair.java

@@ -1,64 +1,0 @@
-package jscheme;
-
-/** A Pair has two fields, first and rest (or car and cdr). 
- * The empty list is represented by null. The methods that you might
- * expect here, like first, second, list, etc. are instead static methods
- * in class SchemeUtils. 
- * @author Peter Norvig, peter@norvig.com http://www.norvig.com 
- * Copyright 1998 Peter Norvig, see http://www.norvig.com/license.html */
-
-public class Pair extends SchemeUtils {
-
-    /** The first element of the pair. **/
-    public Object first;
-
-    /** The other element of the pair. **/
-    public Object rest;
-
-    /** Build a pair from two components. **/
-    public Pair(Object first, Object rest) { 
-	this.first = first; this.rest = rest; 
-    }
-
-    /** Two pairs are equal if their first and rest fields are equal. **/
-    public boolean equals(Object x) {
-	if (x == this) return true;
-	else if (!(x instanceof Pair)) return false;
-	else {
-	  Pair that = (Pair)x;
-	  return equal(this.first, that.first)
-	    && equal(this.rest, that.rest);
-	}
-    }
-
-  /** Return a String representation of the pair. **/
-  public String toString() { return stringify(this, true); }
-
-  /** Build up a String representation of the Pair in a StringBuffer. **/
-  void stringifyPair(boolean quoted, StringBuffer buf) {
-    String special = null;
-    if ((rest instanceof Pair) && rest(rest) == null) 
-      special = (first == "quote") ? "'" : (first == "quasiquote") ? "`"
-	: (first == "unquote") ? "," : (first == "unquote-splicing") ? ",@"
-	: null;
-	
-    if (special != null) {
-      buf.append(special); stringify(second(this), quoted, buf);
-    } else {
-      buf.append('(');
-      stringify(first, quoted, buf);
-      Object tail = rest;
-      while (tail instanceof Pair) {
-	buf.append(' ');
-	stringify(((Pair)tail).first, quoted, buf);
-	tail = ((Pair)tail).rest;
-      }
-      if (tail != null) {
-	buf.append(" . ");
-	stringify(tail, quoted, buf);
-      }
-      buf.append(')');
-    }
-  }
-
-}

### 2010-03-10 src/jscheme/Primitive.java

@@ -1,721 +1,0 @@
-package jscheme;
-import java.io.*;
-
-/** A primitive is a procedure that is defined as part of the Scheme report,
- * and is implemented in Java code. 
- * @author Peter Norvig, peter@norvig.com http://www.norvig.com
- * Copyright 1998 Peter Norvig, see http://www.norvig.com/license.html  **/
-
-public class Primitive extends Procedure {
-
-    int minArgs;
-    int maxArgs;
-    int idNumber;
-
-    public Primitive(int id, int minArgs, int maxArgs) {
-	this.idNumber = id; this.minArgs = minArgs; this.maxArgs = maxArgs;
-    }
-
-    private static final int EQ = 0, LT = 1, GT = 2, GE = 3, LE = 4,
-      ABS = 5, EOF_OBJECT = 6, EQQ = 7, EQUALQ = 8, FORCE = 9,
-      CAR = 10, FLOOR = 11,  CEILING = 12, CONS = 13, 
-      DIVIDE= 14, LENGTH = 15, LIST = 16, LISTQ = 17, APPLY = 18,
-      MAX = 19, MIN = 20, MINUS = 21, NEWLINE = 22, 
-      NOT = 23, NULLQ = 24, NUMBERQ = 25, PAIRQ = 26, PLUS = 27, 
-      PROCEDUREQ = 28, READ = 29, CDR = 30, ROUND = 31, SECOND = 32, 
-      SYMBOLQ = 33, TIMES = 34, TRUNCATE = 35, WRITE = 36, APPEND = 37,
-      BOOLEANQ = 38, SQRT = 39, EXPT = 40, REVERSE = 41, ASSOC = 42, 
-      ASSQ = 43, ASSV = 44, MEMBER = 45, MEMQ = 46, MEMV = 47, EQVQ = 48,
-      LISTREF = 49, LISTTAIL = 50, STRINQ = 51, MAKESTRING = 52, STRING = 53,
-      STRINGLENGTH = 54, STRINGREF = 55, STRINGSET = 56, SUBSTRING = 57, 
-      STRINGAPPEND = 58, STRINGTOLIST = 59, LISTTOSTRING = 60, 
-      SYMBOLTOSTRING = 61, STRINGTOSYMBOL = 62, EXP = 63, LOG = 64, SIN = 65,
-      COS = 66, TAN = 67, ACOS = 68, ASIN = 69, ATAN = 70, 
[…]

### 2010-03-10 src/jscheme/Scheme.java

@@ -1,199 +1,0 @@
-package jscheme;
-import java.io.*;
-
-/** This class represents a Scheme interpreter.
- * See http://www.norvig.com/jscheme.html for more documentation.
- * This is version 1.4. 
- * @author Peter Norvig, peter@norvig.com http://www.norvig.com 
- * Copyright 1998 Peter Norvig, see http://www.norvig.com/license.html **/
-
-public class Scheme extends SchemeUtils {
-
-  InputPort input = new InputPort(System.in);
-  PrintWriter output = new PrintWriter(System.out, true);
-  Environment globalEnvironment = new Environment();
-
-  /** Create a Scheme interpreter and load an array of files into it.
-   * Also load SchemePrimitives.CODE. **/
-  public Scheme(String[] files) { 
-    Primitive.installPrimitives(globalEnvironment); 
-    try {
-      load(new InputPort(new StringReader(SchemePrimitives.CODE)));
-      for (int i = 0; i < (files == null ? 0 : files.length); i++) {
-	load(files[i]);
-      } 
-    } catch (RuntimeException e) { ; }
-  }
-
-  //////////////// Main Loop
-
-  /** Create a new Scheme interpreter, passing in the command line args
-   * as files to load, and then enter a read eval write loop. **/
-  public static void main(String[] files) {
-    new Scheme(files).readEvalWriteLoop();
-  }
-
-  /** Prompt, read, eval, and write the result. 
-   * Also sets up a catch for any RuntimeExceptions encountered. **/
-  public void readEvalWriteLoop() {
-    Object x;
-    for(;;) {
-      try {
-	output.print("> "); output.flush();
-	if (input.isEOF(x = input.read())) return;
-	write(eval(x), output, true); 
-	output.println(); output.flush();
-      } catch (RuntimeException e) { ; }
-    }
-  }
-
-  /** Eval all the expressions in a file. Calls load(InputPort). **/
-  public Object load(Object fileName) {
-    String name = stringify(fileName, false);
-    try { return load(new InputPort(new FileInputStream(name))); }
-    catch (IOException e) { return error("can't load " + name); }
-  }
-
-  /** Eval all the expressions coming from an InputPort. **/
-  public Object load(InputPort in) {
-    Object x = null;
-    for(;;) {
-      if (in.isEOF(x = in.read())) return TRUE;
-      eval(x); 
-    }
-  }
-      
-  //////////////// Evaluation
-
[…]

### 2010-03-10 src/main/Tree.java
Now the jscheme is truly gone.

@@ -1,7 +1,6 @@
 package main;
 
 import java.util.*;
-import jscheme.*;
 
 public class Tree {
     protected Tree parent = null;

### 2010-04-24 src/graphics/TreeDisplay.java
Small changes.  Finally fixed the threading issues (fingers crossed).  AVL Trees may still have one bad rotation operation.

@@ -11,6 +11,7 @@ import java.util.Queue;
 import java.util.Vector;
 import java.util.concurrent.BlockingQueue;
 import java.util.concurrent.LinkedBlockingQueue;
+import java.util.concurrent.Semaphore;
 import java.util.concurrent.TimeUnit;
 
 import javax.swing.JComponent;
@@ -70,7 +71,11 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
 
                     try {
                         Animation a = animations.poll(1, TimeUnit.MICROSECONDS);
-                        if (a != null) a.animate();
+                        while (a != null) {
+                            System.out.println("About to animate " + a);
+                            a.animate();
+                            a = animations.poll(1, TimeUnit.MICROSECONDS);
+                        }
                     } catch (InterruptedException e) {}
                 }
             }
@@ -330,6 +335,30 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
                         repaint();
                         System.out.println("Done animating: " + superTree);
                     }
+
+                    public String toString() { return "animation of " + superTree; }
                 });
+    }
+
+    public void waitForAnimations()
+    {
+        final Semaphore s = new Semaphore(1);
+        try { s.acquire(); } catch (InterruptedException ie) {}
+    	animations.offer(new Animation() {
+                    void animate()
+                    {
+                        System.out.println("About to notify");
+                        s.release();
+                        System.out.println("Notified");
+                    }
+
+                    public String toString() { return "wakeup call"; }
+                });
+        try {
+            System.out.println("Waiting for notification");
+            s.release();
+            System.out.println("Notified");
+            s.acquire();
+        } catch (InterruptedException ie) {}
     }
 }

### 2010-04-24 src/main/Automaton.java

@@ -29,45 +29,33 @@ public class Automaton extends Tree {
         if (kids.size() == 0) {
             if (getTreeName().equals("1")) {
                 setData("q1"); return true;
-            } 
-            if (getTreeName().equals("0")) {
+            } else if (getTreeName().equals("0")) {
                 setData("q0"); return true;
-            } 
-
-            if (getTreeName().equals("q1")) {
+            } else if (getTreeName().equals("q1")) {
                 setData("ACCEPT");
                 return true;
             }
-        }
-
-        if (kids.size() == 1) {
+        } else if (kids.size() == 1) {
             String k = kids.get(0).getTreeName();
             if (!k.equals("q1") && !k.equals("q0"))
                 return kids.get(0).step();
-
-            if (getTreeName().equals("not")) {
+            else if (getTreeName().equals("not")) {
                 removeChildren(false);
                 setData(k.equals("q1") ? "q0" : "q1");
                 return true;
             }
-        }
-
-        if (kids.size() == 2) {
+        } else if (kids.size() == 2) {
             String l = kids.get(0).getTreeName();
             String r = kids.get(1).getTreeName();
             if (!l.equals("q1") && !l.equals("q0")) 
                 return kids.get(0).step();
-
-            if (!r.equals("q1") && !r.equals("q0")) 
+            else if (!r.equals("q1") && !r.equals("q0")) 
                 return kids.get(1).step();
-
-            if (getTreeName().equals("and")) {
+            else if (getTreeName().equals("and")) {
                 removeChildren(false);
                 setData((l.equals("q1") && r.equals("q1")) ? "q1" : "q0");
                 return true;
-            }
-
-            if (getTreeName().equals("or")) {
+            } else if (getTreeName().equals("or")) {
                 removeChildren(false);
                 setData((l.equals("q1") || r.equals("q1")) ? "q1" : "q0");
                 return true;

### 2010-04-24 src/main/MainWindow.java

@@ -38,7 +38,7 @@ public class MainWindow extends JFrame {
             public void run() {
                 while (step()) {
                   try {
-                    Thread.sleep(2000);
+                    Thread.sleep(200);
                   } catch (InterruptedException ie) {}
                 }
             }  
@@ -76,7 +76,9 @@ public class MainWindow extends JFrame {
         startLayout();
     }
 
+    System.out.println("About to step the lisp");
     boolean rv = lisp.step();
+    System.out.println("done with step the lisp");
     outputtext.append("\n" + lisp);
     return rv;
   }

### 2010-04-24 src/main/Tree.java

@@ -23,7 +23,7 @@ public class Tree {
     public Tree(Tree copy)
     {
         setData(copy.getData(), false);
-        for (Tree t : copy.getChildren()) {
+        for (Tree t : new Vector<Tree>(copy.getChildren())) {
             addChild(new Tree(t), false);
         }
     }
@@ -68,7 +68,7 @@ public class Tree {
             this.data = "";
         }
 
-        if (fire) fireDataChangedEvent();
+        if (fire) { fireDataChangedEvent(); waitForAnimations(); }
     }
 
     public String getTreeName()
@@ -101,15 +101,11 @@ public class Tree {
     }
     
     public void setChild(int position, Tree child){
-      synchronized(child){
         children.set(position, child);
-      }
       
-      synchronized (listeners) {
         for (TreeChangeListener l : listeners) {
             child.addTreeChangeListener(l);
         }
-      }
     }
     
 
@@ -132,7 +128,8 @@ public class Tree {
                 child.addTreeChangeListener(l);
             }
         }
-        if (fire) fireTreeAddedEvent(child);
+
+        if (fire) { fireTreeAddedEvent(child); waitForAnimations(); }
     }
 
     public void removeChildren(){ removeChildren(true); }
@@ -149,7 +146,7 @@ public class Tree {
             }
         }
 
-        if (fire) fireChildrenClearedEvent();
+        if (fire) { fireChildrenClearedEvent(); waitForAnimations(); }
     }
 
     int depthcache = -1;
@@ -178,28 +175,20 @@ public class Tree {
     }
 
     public void addTreeChangeListener(TreeChangeListener listener, Set<Tree> heard){
-        synchronized (listeners) {
-            listeners.add(listener);
-        }
+        listeners.add(listener);
 
-        synchronized (children) {
-            for (Tree kid : children) {
-                if (kid == null || heard.contains(kid)) continue;
-                heard.add(kid);
-                kid.addTreeChangeListener(listener, heard);
-            }
+        for (Tree kid : new Vector<Tree>(children)) {
+            if (kid == null || heard.contains(kid)) continue;
+            heard.add(kid);
+            kid.addTreeChangeListener(listener, heard);
         }
     }
 
     public void removeTreeChangeListener(TreeChangeListener listener){
-        synchronized (listeners) {
-            listeners.remove(listener);
-        }
+        listeners.remove(listener);
 
-        synchronized (children) {
-            for (Tree kid : children) {
[…]

### 2010-04-24 src/main/TreeAutomaton.java

@@ -59,6 +59,8 @@ public class TreeAutomaton extends JFrame {
 
       String code = bottomInput.getText();
       code = code.replace("\n", " ");
+      code = code.replace("  ", " ");
+      code = code.trim();
       try {
           automaton = Automaton.parse(code);
           treeDisplay.setTree(automaton);

### 2010-04-24 src/graphics/TreeDisplay.java
Last few unsynchronized threading errors looking done.  Also removed extraneous output.

@@ -72,7 +72,7 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
                     try {
                         Animation a = animations.poll(1, TimeUnit.MICROSECONDS);
                         while (a != null) {
-                            System.out.println("About to animate " + a);
+                            //System.out.println("About to animate " + a);
                             a.animate();
                             a = animations.poll(1, TimeUnit.MICROSECONDS);
                         }
@@ -133,7 +133,7 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
             } 
 
             if (treechanged) {
-                System.out.println("TREE CHANGED");
+                //System.out.println("TREE CHANGED");
                 treechanged = false;
                 layout(tree, 0, PADDING);
             }
@@ -258,7 +258,7 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
     	animations.offer(new Animation() {
                     void animate()
                     {
-                        System.out.println("Animating: " + superTree);
+                        //System.out.println("Animating: " + superTree);
                         // Make the new layout
                         TreeDisplay td = new TreeDisplay(false);
                         td.cachedGraphics = cachedGraphics;
@@ -333,7 +333,7 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
                         }
                         
                         repaint();
-                        System.out.println("Done animating: " + superTree);
+                        //System.out.println("Done animating: " + superTree);
                     }
 
                     public String toString() { return "animation of " + superTree; }
@@ -347,17 +347,17 @@ public class TreeDisplay extends JComponent implements TreeChangeListener{
     	animations.offer(new Animation() {
                     void animate()
                     {
-                        System.out.println("About to notify");
+                        //System.out.println("About to notify");
                         s.release();
-                        System.out.println("Notified");
+                        //System.out.println("Notified");
                     }
 
                     public String toString() { return "wakeup call"; }
                 });
         try {
-            System.out.println("Waiting for notification");
+            //System.out.println("Waiting for notification");
             s.release();
-            System.out.println("Notified");
+            //System.out.println("Notified");
             s.acquire();
         } catch (InterruptedException ie) {}
     }

### 2010-04-24 src/main/BinaryTreeInsertion.java

@@ -72,14 +72,18 @@ public class BinaryTreeInsertion extends LispFreeWindow{
     System.out.println("Bal = " + balance);
     if (balance == -2){ //Right side is too heavy
       if(balanceRatio(root.getChild(RIGHT)) == -1){
+        System.out.println("leftRotation");
         leftRotation(root);
       }else{
+        System.out.println("rightLeft");
         rightLeft(root);
       }
     }else if (balance == 2){
       if (balanceRatio(root.getChild(LEFT)) == 1){
+        System.out.println("rightRotation");
         rightRotation(root);
       }else{
+        System.out.println("leftRight");
         leftRight(root);
       }
     }
@@ -127,7 +131,8 @@ public class BinaryTreeInsertion extends LispFreeWindow{
     c.removeChildren(false);
 
     String adata = a.getData();
-    root.setData(b.getData(), false);
+    a.setData(b.getData(), false);
+    b.setData(adata);
     a = b;
     b = root;
 

### 2010-04-24 src/main/Tree.java

@@ -124,7 +124,7 @@ public class Tree {
         }
 
         synchronized (listeners) {
-            for (TreeChangeListener l : listeners) {
+            for (TreeChangeListener l : new Vector<TreeChangeListener>(listeners)) {
                 child.addTreeChangeListener(l);
             }
         }
@@ -175,7 +175,7 @@ public class Tree {
     }
 
     public void addTreeChangeListener(TreeChangeListener listener, Set<Tree> heard){
-        listeners.add(listener);
+        synchronized (listener) { listeners.add(listener); }
 
         for (Tree kid : new Vector<Tree>(children)) {
             if (kid == null || heard.contains(kid)) continue;
@@ -185,7 +185,7 @@ public class Tree {
     }
 
     public void removeTreeChangeListener(TreeChangeListener listener){
-        listeners.remove(listener);
+        synchronized (listener) { listeners.remove(listener); }
 
         for (Tree kid : new Vector<Tree>(children)) {
             kid.addTreeChangeListener(listener);
@@ -194,7 +194,7 @@ public class Tree {
 
     private void fireTreeAddedEvent(Tree child){
         synchronized (listeners) {
-            for (TreeChangeListener nodeChangeListener : listeners) {
+            for (TreeChangeListener nodeChangeListener : new Vector<TreeChangeListener>(listeners)) {
                 nodeChangeListener.kidAdded(this, child);
             }
         }
@@ -202,7 +202,7 @@ public class Tree {
 
     private void fireChildrenClearedEvent(){
         synchronized (listeners) {
-            for (TreeChangeListener nodeChangeListener : listeners) {
+            for (TreeChangeListener nodeChangeListener : new Vector<TreeChangeListener>(listeners)) {
                 nodeChangeListener.childrenRemoved(this);
             }
         }

### 2010-04-24 rand.py
Automatons work better now.

@@ -1,0 +1,46 @@
+from random import choice, randint
+def rb(): return choice([True, False])
+
+def rand(depth=6, forced=True, t=True):
+    if depth == 0:
+        if forced:
+            if t:
+                return "1"
+            else:
+                return "0"
+        else:
+            if rb():
+                return "1"
+            else:
+                return "0"
+    else:
+        c = choice(["not", "and", "or"])
+        if "not" == c:
+            return "(not " + rand(depth-1, forced, not t) + ")"
+        elif "and" == c:
+            if not forced:
+                return "(and " + rand(depth-1, False, rb()) + " " + rand(depth-1, False, rb()) + ")"
+            elif t:
+                return "(and " + rand(depth-1, forced, True) + " " + rand(depth-1, forced, True) + ")"
+            else:
+                if rb():
+                    return "(and " + rand(depth-1, False, rb()) + " " + rand(depth-1, True, False) + ")"
+                else:
+                    return "(and " + rand(depth-1, True, False) + " " + rand(depth-1, False, rb()) + ")"
+        elif "or" == c:
+            if not forced:
+                return "(or " + rand(depth-1, forced, rb()) + " " + rand(depth-1, forced, rb()) + ")"
+            elif not t:
+                return "(or " + rand(depth-1, forced, False) + " " + rand(depth-1, forced, False) + ")"
+            else:
+                if rb():
+                    return "(or " + rand(depth-1, False, rb()) + " " + rand(depth-1, True, True) + ")"
+                else:
+                    return "(or " + rand(depth-1, True, True) + " " + rand(depth-1, False, rb()) + ")"
+        
+            
+    
+    
+
+if __name__ == '__main__':
+    print rand()

### 2010-04-24 src/main/Tree.java

@@ -103,8 +103,10 @@ public class Tree {
     public void setChild(int position, Tree child){
         children.set(position, child);
       
-        for (TreeChangeListener l : listeners) {
-            child.addTreeChangeListener(l);
+        synchronized (listeners) {
+            for (TreeChangeListener l : listeners) {
+                child.addTreeChangeListener(l);
+            }
         }
     }
     
@@ -218,7 +220,12 @@ public class Tree {
 
     private void waitForAnimations()
     {
-        for (TreeChangeListener nodeChangeListener : new Vector<TreeChangeListener>(listeners)) {
+        Vector<TreeChangeListener> l;
+        synchronized (listeners) {
+            l = new Vector<TreeChangeListener>(listeners);
+        }
+
+        for (TreeChangeListener nodeChangeListener : l) {
             nodeChangeListener.waitForAnimations();
         }
     }

### 2010-04-24 src/main/TreeAutomaton.java

@@ -103,7 +103,16 @@ public class TreeAutomaton extends JFrame {
             } catch (IOException ioe){
                 System.err.println("Error reading: " + args[i]);
             }
-       }
+       } 
+    }
+    if (args.length == 0) {
+        m.bottomInput.text.append(
+                    "(and (not (and (not (or (or 1 1) (not 1))) (or (and (not 1) (and 0\n"
+                    + "0)) (or (or 1 1) (or 1 0))))) (not (and (and (or (not 1) (and 0 1))\n"
+                    + "(and (or 0 0) (or 1 1))) (or (or (and 1 0) (and 0 1)) (or (and 1\n"
+                    + "1) (or 0 0))))))"
+                );
+
     }
   }
 }

### 2010-04-25 rand.py
Now has a default boolean statement, and uses "true" and "false" instead of "q1" and "q0"

@@ -1,46 +1,32 @@
 from random import choice, randint
 def rb(): return choice([True, False])
 
-def rand(depth=6, forced=True, t=True):
+def rand(depth=6, t=True):
     if depth == 0:
-        if forced:
-            if t:
-                return "1"
-            else:
-                return "0"
+        if t:
+            return "1"
         else:
-            if rb():
-                return "1"
-            else:
-                return "0"
+            return "0"
     else:
         c = choice(["not", "and", "or"])
         if "not" == c:
-            return "(not " + rand(depth-1, forced, not t) + ")"
+            return "(not " + rand(depth-1, not t) + ")"
         elif "and" == c:
-            if not forced:
-                return "(and " + rand(depth-1, False, rb()) + " " + rand(depth-1, False, rb()) + ")"
-            elif t:
-                return "(and " + rand(depth-1, forced, True) + " " + rand(depth-1, forced, True) + ")"
+            if t:
+                return "(and " + rand(depth-1, True) + " " + rand(depth-1, True) + ")"
             else:
                 if rb():
-                    return "(and " + rand(depth-1, False, rb()) + " " + rand(depth-1, True, False) + ")"
+                    return "(and " + rand(depth-1, rb()) + " " + rand(depth-1, False) + ")"
                 else:
-                    return "(and " + rand(depth-1, True, False) + " " + rand(depth-1, False, rb()) + ")"
+                    return "(and " + rand(depth-1, False) + " " + rand(depth-1, rb()) + ")"
         elif "or" == c:
-            if not forced:
-                return "(or " + rand(depth-1, forced, rb()) + " " + rand(depth-1, forced, rb()) + ")"
-            elif not t:
-                return "(or " + rand(depth-1, forced, False) + " " + rand(depth-1, forced, False) + ")"
+            if not t:
+                return "(or " + rand(depth-1, False) + " " + rand(depth-1, False) + ")"
             else:
                 if rb():
-                    return "(or " + rand(depth-1, False, rb()) + " " + rand(depth-1, True, True) + ")"
+                    return "(or " + rand(depth-1, rb()) + " " + rand(depth-1, True) + ")"
                 else:
-                    return "(or " + rand(depth-1, True, True) + " " + rand(depth-1, False, rb()) + ")"
+                    return "(or " + rand(depth-1, True) + " " + rand(depth-1, rb()) + ")"
         
-            
-    
-    
-
 if __name__ == '__main__':
     print rand()

### 2010-04-25 src/main/Automaton.java

@@ -28,36 +28,36 @@ public class Automaton extends Tree {
         // Leaf case
         if (kids.size() == 0) {
             if (getTreeName().equals("1")) {
-                setData("q1"); return true;
+                setData("true"); return true;
             } else if (getTreeName().equals("0")) {
-                setData("q0"); return true;
-            } else if (getTreeName().equals("q1")) {
+                setData("false"); return true;
+            } else if (getTreeName().equals("true")) {
                 setData("ACCEPT");
                 return true;
             }
         } else if (kids.size() == 1) {
             String k = kids.get(0).getTreeName();
-            if (!k.equals("q1") && !k.equals("q0"))
+            if (!k.equals("true") && !k.equals("false"))
                 return kids.get(0).step();
             else if (getTreeName().equals("not")) {
                 removeChildren(false);
-                setData(k.equals("q1") ? "q0" : "q1");
+                setData(k.equals("true") ? "false" : "true");
                 return true;
             }
         } else if (kids.size() == 2) {
             String l = kids.get(0).getTreeName();
             String r = kids.get(1).getTreeName();
-            if (!l.equals("q1") && !l.equals("q0")) 
+            if (!l.equals("true") && !l.equals("false")) 
                 return kids.get(0).step();
-            else if (!r.equals("q1") && !r.equals("q0")) 
+            else if (!r.equals("true") && !r.equals("false")) 
                 return kids.get(1).step();
             else if (getTreeName().equals("and")) {
                 removeChildren(false);
-                setData((l.equals("q1") && r.equals("q1")) ? "q1" : "q0");
+                setData((l.equals("true") && r.equals("true")) ? "true" : "false");
                 return true;
             } else if (getTreeName().equals("or")) {
                 removeChildren(false);
-                setData((l.equals("q1") || r.equals("q1")) ? "q1" : "q0");
+                setData((l.equals("true") || r.equals("true")) ? "true" : "false");
                 return true;
             }
         }

### 2010-04-27 rand.py
New pics and formatting cleanup.

@@ -1,7 +1,7 @@
 from random import choice, randint
 def rb(): return choice([True, False])
 
-def rand(depth=6, t=True):
+def rand(depth=4, t=True):
     if depth == 0:
         if t:
             return "1"

### 2010-04-27 src/graphics/TreeDisplay.java

@@ -3,7 +3,11 @@ package graphics;
 import java.awt.Color;
 import java.awt.Graphics;
 import java.awt.Graphics2D;
+import java.awt.Image;
+import java.awt.image.RenderedImage;
 import java.awt.RenderingHints;
+import java.io.File;
+import java.io.IOException;
 import java.util.HashMap;
 import java.util.LinkedList;
 import java.util.Map;
@@ -15,76 +19,70 @@ import java.util.concurrent.Semaphore;
 import java.util.concurrent.TimeUnit;
 
 import javax.swing.JComponent;
+import javax.imageio.*;
 
 import main.Tree;
 import main.TreeChangeListener;
 
-public class TreeDisplay extends JComponent implements TreeChangeListener{
+public class TreeDisplay extends JComponent implements TreeChangeListener {
 
     private static final Color BACKGROUND = Color.decode("0xAAffff");
     private static final int PADDING = 20;
-
     protected Map<Tree, Node> positions;
     protected Vector<Vector<Node>> layers;
-
     //This is probably over kill, Dr, Boothe gets to decide.
     private BlockingQueue<Animation> animations = new LinkedBlockingQueue<Animation>();
-
     private Tree tree;
     private String message = "Enter some code to run!";
     private boolean treechanged = false;
     private boolean laidout = false;
 
-    public TreeDisplay(boolean shouldgo) 
-    {
-    	super();
-    	if (shouldgo) start();
+    public TreeDisplay(boolean shouldgo) {
+        super();
+        if (shouldgo) {
+            start();
+        }
     }
-    
-    public TreeDisplay()
-    {
+
+    public TreeDisplay() {
         this(true);
     }
 
-    public void setTree(Tree tree)
-    {
+    public void setTree(Tree tree) {
         this.tree = tree;
         treechanged = true;
         message = null;
         laidout = false;
     }
 
-    public void setMessage(String message) 
-    {
+    public void setMessage(String message) {
         this.message = message;
     }
 
-    public void start()
-    {
+    public void start() {
         new Thread() {
-            public void run()
-            {
+            @Override
+            public void run() {
                 while (true) {
                     adjust();
                     repaint();
-                    try { Thread.sleep(1); } catch (InterruptedException ie) {}
 
                     try {
                         Animation a = animations.poll(1, TimeUnit.MICROSECONDS);
                         while (a != null) {
-                            //System.out.println("About to animate " + a);
+                            System.out.println("About to animate " + a);
[…]

### 2010-04-27 src/main/BinaryTreeInsertion.java

@@ -7,217 +7,216 @@ import java.util.Vector;
 
 import javax.swing.JOptionPane;
 
+public class BinaryTreeInsertion extends LispFreeWindow {
 
-public class BinaryTreeInsertion extends LispFreeWindow{
-  
-  final int LEFT = 0;
-  final int RIGHT = 1;
-  
-  public BinaryTreeInsertion(){
-    super("Binary Tree Insertion");
+    final int LEFT = 0;
+    final int RIGHT = 1;
 
-    bottomInput.setPlayAllButtonAction(new ActionListener() {
-            public void actionPerformed(ActionEvent e){ 
+    public BinaryTreeInsertion() {
+        super("Binary Tree Insertion");
+
+        bottomInput.setPlayAllButtonAction(new ActionListener() {
+
+            public void actionPerformed(ActionEvent e) {
                 treeDisplay.setTree(tree = getStartingTree());
             }
         });
-  }
-  
-  protected Tree getStartingTree(){
-    return new Tree("", null);
-  }
-  
-  protected void step(){
-    if (bottomInput.getText().isEmpty() == false){
-      System.out.println(bottomInput.getText());
-      String[] numbers = bottomInput.getText().split(",");
-      try{
-          for(String num : numbers){
-              insertNumber(Integer.parseInt(num.trim()), tree);
-                
-              //try { Thread.sleep(2000); } catch (InterruptedException ie) {}
-          }
-
-          tree.setData(tree.getData());
-
-          bottomInput.text.setText("");
-      }catch(NumberFormatException nfe){
-          JOptionPane.showMessageDialog(this, "Please enter a list of comma seperated numbers");
-      }
     }
-  }
-  
-  private void insertNumber(int num, Tree t){
-        
-    System.out.println("adding num: " + num + " to " + t);
-    if (t.getData() == null || t.getData().isEmpty()){
-      t.setData(num, false);
-      t.addChild(new Tree("", null), false);
-      t.addChild(new Tree("", null));
-    }else if ( num  < Integer.parseInt(t.getData())){
-      insertNumber(num, t.getChild(LEFT));
-      System.out.println("About to rebalance " + t);
-      rebalance(t);
-      System.out.println("done rebalancing " + t);
-    }else{
-      insertNumber(num, t.getChild(RIGHT));
-      System.out.println("About to rebalance " + t);
[…]

### 2010-04-27 src/main/Tree.java

@@ -3,25 +3,30 @@ package main;
 import java.util.*;
 
 public class Tree {
+
     protected Tree parent = null;
     private String data = null;
     private Vector<Tree> children = new Vector<Tree>();
-    
-    Set<TreeChangeListener> listeners = new HashSet<TreeChangeListener>();
+    final Set<TreeChangeListener> listeners = new HashSet<TreeChangeListener>();
 
-    public Tree() {}
-    public Tree(Object data, Iterable<Tree> kids) 
-    {
+    public Tree() {
+    }
+
+    public Tree(Object data, Iterable<Tree> kids) {
         setData(data, false);
-        if (kids != null) 
+        if (kids != null) {
             for (Tree kid : kids) {
                 //Avoid setting off the listeners
                 addChild(kid, false);
             }
+        }
     }
-    public Tree(Iterable<Tree> kids) { this(null, kids); }
-    public Tree(Tree copy)
-    {
+
+    public Tree(Iterable<Tree> kids) {
+        this(null, kids);
+    }
+
+    public Tree(Tree copy) {
         setData(copy.getData(), false);
         for (Tree t : new Vector<Tree>(copy.getChildren())) {
             addChild(new Tree(t), false);
@@ -33,87 +38,94 @@ public class Tree {
     }
 
     @Override
-    public String toString()
-    {
-        if (getData() != null && getChildren().size() == 0)
+    public String toString() {
+        if (getData() != null && getChildren().size() == 0) {
             return getData();
+        }
 
         String rv = "";
         for (Tree kid : getChildren()) {
-            if (kid == this)
-              continue;
+            if (kid == this) {
+                continue;
+            }
             rv += " " + kid;
         }
 
-        if (getData() == null)
+        if (getData() == null) {
             return "(" + rv.trim() + ")";
-        else
+        } else {
             return "(" + getData() + " " + rv.trim() + ")";
+        }
     }
 
     /**
      * @return the data
      */
-    public String getData() 
-    {
+    public String getData() {
         return data;
     }
 
-    protected void setData(Object o) { setData(o, true); }
-    protected void setData(Object o, boolean fire)
-    {
+    protected void setData(Object o) {
+        setData(o, true);
+    }
+
+    protected void setData(Object o, boolean fire) {
         if (o != null) {
             this.data = o.toString();
         } else {
             this.data = "";
         }
 
-        if (fire) { fireDataChangedEvent(); waitForAnimations(); }
+        if (fire) {
+            fireDataChangedEvent();
+            waitForAnimations();
+        }
[…]

### 2010-04-28 src/main/Lisp.java
Added ^ and made the numbers Doubles

@@ -6,7 +6,7 @@ import java.util.*;
 public class Lisp {
   private static final String FUN = "fun";
   private static final String IF = "if";
-  String reservedwords[] = { "+", "-", "*", "/", "<", ">", "=", IF, FUN};
+  String reservedwords[] = { "+", "-", "*", "/", "<", ">", "=", "^", IF, FUN};
 
   Tree root = new Tree();
   public Lisp(String code) throws CompilationException
@@ -228,8 +228,8 @@ public class Lisp {
 
     if (code.getData().equals("<")) {
         Vector<Tree> ltargs = code.getChildren();
-        Integer left = Integer.parseInt(ltargs.get(0).getData());
-        Integer right = Integer.parseInt(ltargs.get(1).getData());
+        Double left = Double.parseDouble(ltargs.get(0).getData());
+        Double right = Double.parseDouble(ltargs.get(1).getData());
 
         code.removeChildren(false);
         if (left < right) {
@@ -241,8 +241,8 @@ public class Lisp {
         return true;
     } else if (code.getData().equals(">")) {
         Vector<Tree> ltargs = code.getChildren();
-        Integer left = Integer.parseInt(ltargs.get(0).getData());
-        Integer right = Integer.parseInt(ltargs.get(1).getData());
+        Double left = Double.parseDouble(ltargs.get(0).getData());
+        Double right = Double.parseDouble(ltargs.get(1).getData());
 
         code.removeChildren(false);
         if (left > right) {
@@ -254,8 +254,8 @@ public class Lisp {
         return true;
     } else if (code.getData().equals("=")) {
         Vector<Tree> ltargs = code.getChildren();
-        Integer left = Integer.parseInt(ltargs.get(0).getData());
-        Integer right = Integer.parseInt(ltargs.get(1).getData());
+        Double left = Double.parseDouble(ltargs.get(0).getData());
+        Double right = Double.parseDouble(ltargs.get(1).getData());
 
         code.removeChildren(false);
         if (left.equals(right)) {
@@ -267,8 +267,8 @@ public class Lisp {
         return true;
     } else if (code.getData().equals("+")) {
         Vector<Tree> ltargs = code.getChildren();
-        Integer left = Integer.parseInt(ltargs.get(0).getData());
-        Integer right = Integer.parseInt(ltargs.get(1).getData());
+        Double left = Double.parseDouble(ltargs.get(0).getData());
+        Double right = Double.parseDouble(ltargs.get(1).getData());
 
         code.removeChildren(false);
[…]

### 2010-04-30 paper/paper.tex
Starting in on the paper

\documentclass{article}

\usepackage{url}

\title{Drawing Trees and Animating Tree Changes}

\author{Sandro Badame, Peter Boothe$^*$\\
 Manhattan College\\
\url{{sbadame.student,peter.boothe}@manhattan.edu}}

\begin{document}
\maketitle

\begin{abstract}
summary.  poster verbage
\end{abstract}

\end{document}

### 2010-04-30 paper/paper.tex
4 pages, looking good.  Needs lots more meat, but it's coming along.

@@
 \usepackage{url}
+\usepackage{graphicx}
 
@@
 
+\newcommand{\newword}[1]{\textit{#1}\index{#1}}
+\newcommand{\java}[1]{\texttt{#1}\index{#1}}
+
 \begin{document}
@@
 \begin{abstract}
-summary.  poster verbage
-\end{abstract}
+Dynamic trees appear in many areas of computer science, including, but not
+limited to, Lisp code, data structures, tree automata, and computer
+algorithms.  It would be useful to be able to easily draw trees and animate
+them changing over time.  We begin by articulating a design for tree
+visualization and animation that is more general than previous methodologies.
+We demonstrate, using several visualizations, the
+usage of a library we have created which attempts to simplify the graphics
+aspect of this tricky task.\end{abstract}
+
+\section{Introduction}
+Trees, and especially trees which change over time, are fundamental objects in
+computer science.  They occur in many disparate contexts, and while the uses of
+these trees may be different, there are some commonalities which we can use to
+build a tree visualization system which might be useful in many different
+contexts.
+In this paper we articulate the design principles of a general tree visualization system, present one such system that we have built in Java and made freely available, and then demonstrate our system's utility with several sample visualizations.
+
+There are multiple reasons to visualize a system, and different reasons have
+different requirements for the visualization.  A visualization designed for
+experts may be very useful for that target group, but leave neophytes
+confused.  Therefore, before we discuss the particulars of our visualization,
+we must explain our reasons for wanting such a system in the first place, and
+the expected audience.  We want to visualize dynamic trees for demonstration,
+education, and beauty, and we would like our visualizations to be of benefit to
+computer science undergraduates.
+
+\subsection{Visualization for Demonstration}
+
+We want to show them how things work.  We therefore will restrict ourselves to trees which are small enough to be drawn in their entirety on a single computer screen.
+
+\subsection{Visualization for Education}
+
+It turns out that demonstrations are often useless.  But if we force students to build their own visualizations, then they actually learn.  So We would like to enable that as well.
+
+\subsection{Visualization for Beauty}
+
+Computational processes are often both sophisticated and beautiful, but their
+beauty is often lost in the nitty-gritty of simply getting a given algorithm to
+work.  Using a visualization system could, hopefully, remind students of the
+beauty of the structures they are working with and also serve as an effective
+advertisement to non-majors of the intricacy and interestingness available in
+the CS major.
+
+\section{Designing a Tree Visualization Framework}
+
+Let us begin our design by formally defining the objects under consideration.
+A \newword{tree} $T$ is an undirected connected acyclic graph with vertex set
+$V$ and edge set $E \subseteq \{V \times V\}$.  A \newword{rooted tree} is a
+tree where a particular vertex $r$ is designated the as the \newword{root} of
+the tree, which allows us to define the \newword{depth} of a vertex $v$ as the
+length of the shortest path between $v$ and $r$. Every edge connects two
+vertices $(u,v)$, and we will assume, again wlog, that the depth of $u$ is one
+less than the depth of $v$.  In that case, we call $u$ the \newword{parent} of
+$v$ and $v$ the \newword{child} of $u$.  The \newword{children} of $v$ are the
+set of vertices which have $v$ as their parent.  A \newword{rooted ordered
+tree} is a rooted tree with a partial-ordering on the vertex set $V$ such that
+for all vertices $\{u,v, \ldots\}$ with the same parent, either $u < v$ or $v <
+u$.
+
+Without loss of generalization, we restrict our visualization system to rooted,
+ordered trees.  Any tree can be converted into a rooted, ordered tree by
+selecting a root and imposing a partial-ordering on the vertices.  In practice,
+however, we found that every domain we considered had a naturally occuring root
+and ordering, and that the imposition of an arbitrary root and ordering was
+unnecessary.  From now on, when we say tree, we will mean a rooted ordered tree.
+
+To lay out a tree in the plane, we must map every vertex $v \in V$ to a
+coordinate $(v_x, v_y)$.  When visualizing a tree we put the root at the top
+(which, while counter-intuitive for biologists, is standard practice in
+computer science), and every vertex $\{u,v \ldots\} \subset V$ with equal depth
+should be laid out with the same $y$-coordinate $(u_y = v_y)$.  Furthermore, if
+two nodes $u$ and $v$ have the same parent and $u < v$, then we should lay them
+out with $u_x < v_x$, which will preserve in our layout the natural order of the tree.
+Our last requirement is that if we have two nodes $w$ and $z$ with equal depth and children $U$ and $V$, and $w_x < z_x$, then $u_x < v_x \forall u\in U, v\in V$.
+This last requirement allows us to avoid crossing edges.
+
+\subsection{Prior Work}
+
+Paper from 1996 and Tamassia and FruchetermanWhatever magnets and springs
+model.
+
+\subsection{Taking Advantage of Computational Abundance}
+
+Past visualization frameworks went to great lengths to articulate the allowed
+set of transformations which allowed them to heavily optimize their layout code
+to minimize CPU usage.  In our situation, we would like to make our library as
+usable as possible.  Therefore, we take every possible opportunity to make our
+library easy to use, even if it increases the polynomial runtime of some of our
+methods.  This is safe to do because we have restricted ourself to those graphs
+which can be drawn, in their entirety, on a computer screen, and have therefore
+restricted ourself to the (admittedly ill-defined) set of graphs which are
+``small enough'' to still be drawn quickly using algorithms which may be of
+order $\Omega(n^2)$, as long as these algorithms are not super-polynomial.
+
+Therefore, instead of restricting ourself to a particular subset of allowed
+tree operations, we allow arbitrary transformations of the tree, and then we
+animate our old tree transforming into the new tree.
+
+\subsection{Layout Algorithm}
+
+1) Fix the root node to the middle top position. 
+2) For each node of a given level treat every 
+link as a spring and every node as a magnet 
+and calculate the resultant forces. 
+3) Simulate the change in position due to 
+these forces and update all node positions 
+by a small amount. 
+4) Display the system to the user, and repeat
+
+\subsection{Animation Algorithm}
+
+Either on every tree transformation, or when requested by the visualization programmer, we relayout the entire tree according to our layout algorithm, and then we animate the transition from the older, currently-laid-out tree, to the new tree.
+Or animation proceeds in the following steps:
+
+1) Layout the new tree using the method above. 
+Do this layout off screen. 
+2) Fade out any of the nodes of the currently 
+displayed original tree that have been removed. 
+3) Compare the locations of the nodes in the 
+original tree to those in the new tree, then 
+animate the movement of the original nodes 
+to their new positions. 
+4) Fade in any nodes that only exist in the 
+new tree. 
+Multiple changes can be either combined into a 
+single animation step, or each step may be 
+animated individually.  This allows artistic flexibility 
+when creating a new visualization.
+
+\section{Our Library}
+
+Our library is written in Java, and users of the library need only use two of the classes it contains: the \java{Tree} class, and the \java{TreeDisplay} class.
+
+\subsection{How-To}
+\subsection{A Sample Application}
+
+\section{Example Animations}
+\subsection{Random Tree Growth}
+\subsection{Self-Balancing Binary Search Tree}
+\subsection{Tree Automaton}
+\begin{figure}
+\begin{center}
+\includegraphics[width=1in]{../board/pics/ss301.png}
+\includegraphics[width=1in]{../board/pics/ss351.png}
+\includegraphics[width=1in]{../board/pics/ss401.png}
+\includegraphics[width=1in]{../board/pics/ss451.png}
+\includegraphics[width=1in]{../board/pics/ss501.png}
+\includegraphics[width=1in]{../board/pics/ss551.png}
+\includegraphics[width=1in]{../board/pics/ss601.png}
+\includegraphics[width=1in]{../board/pics/ss651.png}
+\includegraphics[width=1in]{../board/pics/ss701.png}
+\includegraphics[width=1in]{../board/pics/ss751.png}
+\includegraphics[width=1in]{../board/pics/ss801.png}
+\includegraphics[width=1in]{../board/pics/ss851.png}
+\includegraphics[width=1in]{../board/pics/ss901.png}
+\includegraphics[width=1in]{../board/pics/ss951.png}
+\includegraphics[width=1in]{../board/pics/ss1001.png}
+\includegraphics[width=1in]{../board/pics/ss1051.png}
+\includegraphics[width=1in]{../board/pics/ss1101.png}
+\includegraphics[width=1in]{../board/pics/ss1151.png}
+\includegraphics[width=1in]{../board/pics/ss1201.png}
+\includegraphics[width=1in]{../board/pics/ss1251.png}
+\includegraphics[width=1in]{../board/pics/ss1301.png}
+\includegraphics[width=1in]{../board/pics/ss1351.png}
+\includegraphics[width=1in]{../board/pics/ss1401.png}
+\includegraphics[width=1in]{../board/pics/ss1451.png}
+\includegraphics[width=1in]{../board/pics/ss1501.png}
+\includegraphics[width=1in]{../board/pics/ss1551.png}
+\includegraphics[width=1in]{../board/pics/ss1601.png}
+\includegraphics[width=1in]{../board/pics/ss1651.png}
+\includegraphics[width=1in]{../board/pics/ss1701.png}
+\includegraphics[width=1in]{../board/pics/ss1751.png}
+\includegraphics[width=1in]{../board/pics/ss1801.png}
+\end{center}
+\caption{Execution of a tree automaton which recognizes tautologies}
+\end{figure}
+
+\subsection{Lisp}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=1.25in]{../board/pics/qf1101.png}
+\includegraphics[width=1.25in]{../board/pics/qf1551.png}
+\includegraphics[width=1.25in]{../board/pics/qf1601.png}
+\includegraphics[width=1.25in]{../board/pics/qf1651.png}
+\includegraphics[width=1.25in]{../board/pics/qf1701.png}
+\includegraphics[width=1.25in]{../board/pics/qf1801.png}
+\includegraphics[width=1.25in]{../board/pics/qf1951.png}
+\includegraphics[width=1.25in]{../board/pics/qf2001.png}
+\includegraphics[width=1.25in]{../board/pics/qf2101.png}
+\includegraphics[width=1.25in]{../board/pics/qf2151.png}
+\includegraphics[width=1.25in]{../board/pics/qf2251.png}
+\includegraphics[width=1.25in]{../board/pics/qf2301.png}
+\includegraphics[width=1.25in]{../board/pics/qf2351.png}
+\includegraphics[width=1.25in]{../board/pics/qf2401.png}
+\includegraphics[width=1.25in]{../board/pics/qf2551.png}
+\includegraphics[width=1.25in]{../board/pics/qf2601.png}
+\includegraphics[width=1.25in]{../board/pics/qf2701.png}
+\includegraphics[width=1.25in]{../board/pics/qf2801.png}
+\end{center}
+\caption{The complete execution of a lisp program calculating one of the roots of $x^2+x-6$.}
+\end{figure}
+
+\section{Conclusion}
+
+We have passed a threshold of computational abundance which can allow visualization libraries to use less-efficient algorithms in pursuit of programming ease.
+Noting this threshold allows us to articulate a design for a layout library that, by doing much more work behind the scenes, makes the creation and usage of visualizations available to a much larger audience.
+We have created such a library and provided several sample visualizations.
+We hope that any and all of our design ideas, the resulting library, and the provided visualizations, might be useful to others who are attempting to demonstrate, learn, understand, or admire the behavior of trees which evolve over time.
 

### 2010-05-12 paper/paper.tex
Small paper changes.

@@
 \usepackage{graphicx}
+\usepackage[width=4.5in,height=7.2in]{geometry}
 
@@
 
-\author{Sandro Badame, Peter Boothe$^*$\\
+\author{Sandro Badame, Peter Boothe\footnote{Corresponding author}\\
  Manhattan College\\
@@
 computer science.  They occur in many disparate contexts, and while the uses of
-these trees may be different, there are some commonalities which we can use to
+these trees may be different, there exist commonalities which we can use to
 build a tree visualization system which might be useful in many different
@@
 There are multiple reasons to visualize a system, and different reasons have
-different requirements for the visualization.  A visualization designed for
-experts may be very useful for that target group, but leave neophytes
-confused.  Therefore, before we discuss the particulars of our visualization,
-we must explain our reasons for wanting such a system in the first place, and
-the expected audience.  We want to visualize dynamic trees for demonstration,
-education, and beauty, and we would like our visualizations to be of benefit to
-computer science undergraduates.
+different requirements.  A visualization designed for experts may be very
+useful for that target group, but leave neophytes confused.  Therefore, before
+we discuss the particulars of our visualization, we must explain our reasons
+for wanting such a system in the first place, and the expected audience.  We
+want to visualize dynamic trees for demonstration, education, and beauty, and
+we would like our visualizations to be of benefit to computer science
+undergraduates.
 
@@
 
-It turns out that demonstrations are often useless.  But if we force students to build their own visualizations, then they actually learn.  So We would like to enable that as well.
+It turns out that demonstrations are often useless.  But if we force students to build their own visualizations, then they actually learn.  So we would like to enable that as well.
 
@@
 ``small enough'' to still be drawn quickly using algorithms which may be of
-order $\Omega(n^2)$, as long as these algorithms are not super-polynomial.
+order $\omega(n)$, as long as these algorithms are not super-polynomial.
 

### 2010-05-14 paper/paper.tex
Small changes, needs merging
Lots of reflowing of text, but nothing major.

@@
 Dynamic trees appear in many areas of computer science, including, but not
-limited to, Lisp code, data structures, tree automata, and computer
-algorithms.  It would be useful to be able to easily draw trees and animate
-them changing over time.  We begin by articulating a design for tree
-visualization and animation that is more general than previous methodologies.
-We demonstrate, using several visualizations, the
-usage of a library we have created which attempts to simplify the graphics
-aspect of this tricky task.\end{abstract}
+limited to, Lisp code, data structures, tree automata, and computer algorithms.
+It would be useful to be able to easily draw trees and animate them changing
+over time.  We begin by articulating a design for tree visualization and
+animation that is more general than previous methodologies.  We demonstrate,
+using several visualizations, the usage of a library we have created which
+attempts to simplify the graphics aspect of this tricky task.\end{abstract}
 
@@
 
-Paper from 1996 and Tamassia and FruchetermanWhatever magnets and springs
-model.
+Paper from 1996 and Tamassia and the force-directed placement algorithm\cite{fdp}.
 
@@
 
-Either on every tree transformation, or when requested by the visualization programmer, we relayout the entire tree according to our layout algorithm, and then we animate the transition from the older, currently-laid-out tree, to the new tree.
-Or animation proceeds in the following steps:
+Either on every tree transformation, or when requested by the visualization
+programmer, we relayout the entire tree according to our layout algorithm, and
+then we animate the transition from the older, currently-laid-out tree, to the
+new tree.  Or animation proceeds in the following steps:
 
@@
 
-Our library is written in Java, and users of the library need only use two of the classes it contains: the \java{Tree} class, and the \java{TreeDisplay} class.
+Our library is written in Java, and users of the library need only use two of
+the classes it contains: the \java{Tree} class, and the \java{TreeDisplay}
+class.  Our primary design principle in building this system was ease of usage
+by another programmer.  Therefore, our two classes have very few methods, and
+those few methods are very simple to use.  Prior efforts focused a lot of work
+on making specific animations efficient, but we have instead focused our
+efforts on ease of use.
 
 \subsection{How-To}
+
+The class \java{TreeDisplay} needs to only be created once.  A
+\java{TreeDisplay} is an AWT component, and may be used as such in any Java
+code.  A \java{Tree} may be created at any time, and to display a tree, use the
+\java{setTree} method of the \java{TreeDisplay} class.
+
 \subsection{A Sample Application}
@@
 
-We have passed a threshold of computational abundance which can allow visualization libraries to use less-efficient algorithms in pursuit of programming ease.
-Noting this threshold allows us to articulate a design for a layout library that, by doing much more work behind the scenes, makes the creation and usage of visualizations available to a much larger audience.
-We have created such a library and provided several sample visualizations.
-We hope that any and all of our design ideas, the resulting library, and the provided visualizations, might be useful to others who are attempting to demonstrate, learn, understand, or admire the behavior of trees which evolve over time.
+We have passed a threshold of computational abundance which can allow
+visualization libraries to use less-efficient algorithms in pursuit of
+programming ease.  Noting this threshold allows us to articulate a design for a
+layout library that, by doing much more work behind the scenes, makes the
+creation and usage of visualizations available to a much larger audience.  We
+have created such a library and provided several sample visualizations.  We
+hope that any and all of our design ideas, the resulting library, and the
+provided visualizations, might be useful to others who are attempting to
+demonstrate, learn, understand, or admire the behavior of trees which evolve
+over time.
+
+\bibliographystyle{plain}
+\bibliography{bibliography.bib}
 

### 2010-05-14 paper/paper.tex
Almost go for launch

@@
 
-\begin{abstract}
-Dynamic trees appear in many areas of computer science, including, but not
-limited to, Lisp code, data structures, tree automata, and computer algorithms.
-It would be useful to be able to easily draw trees and animate them changing
-over time.  We begin by articulating a design for tree visualization and
-animation that is more general than previous methodologies.  We demonstrate,
-using several visualizations, the usage of a library we have created which
-attempts to simplify the graphics aspect of this tricky task.\end{abstract}
+\begin{abstract} Dynamic trees appear in many areas of computer science,
+including, but not limited to, Lisp code, data structures, tree automata, and
+computer algorithms.  It would be useful to be able to easily draw trees and
+animate them changing over time.  We begin by articulating a design for tree
+visualization and animation that is more general than previous methodologies.
+Our design is enabled by the observation that, for trees which are small enough
+to be displayed on a screen, computers are now fast enough to eliminate many
+worries about efficiency.  This allows us to make a library which emphasizes
+simplicity of use over efficiency of execution.  We demonstrate our library
+using several visualizations.\end{abstract}
 
@@
 build a tree visualization system which might be useful in many different
-contexts.
-In this paper we articulate the design principles of a general tree visualization system, present one such system that we have built in Java and made freely available, and then demonstrate our system's utility with several sample visualizations.
+contexts.  In this paper we articulate the design principles of a highly usable
+tree visualization system, present one such system that we have built in Java
+and made freely available, and then demonstrate our system's utility with
+several sample visualizations.
 
@@
 
-We want to show them how things work.  We therefore will restrict ourselves to trees which are small enough to be drawn in their entirety on a single computer screen.
+One reason for visualization is for demonstration.  Trees (and other systems)
+are visualized in an effort to show how these systems work.  Because this is
+one of our goals, we will restrict ourselves to trees which are small enough to
+be drawn in their entirety on a single computer screen.
+Whenever we visualize a tree, our display system will show the entire tree, and it will be part of the job of our library to make the best use of the available screen real estate.
 
 \subsection{Visualization for Education}
+\label{educ}
 
-It turns out that demonstrations are often useless.  But if we force students to build their own visualizations, then they actually learn.  So we would like to enable that as well.
+Computing and math eductors have a long history of building visualization
+systems for demonstration, on the grounds that if students can view these
+systems in action, then the students will understand these system.
+Unfortunately for those goals, Hundhausen, Douglas, and Stasko found that, in practice, almost all
+visualizations are useless\cite{hds}.  A good visualization seems intuitively useful, but
+it turns out that if a student does not understand the system being visualized, they will not understand the visualization either.
+The one bright spot in this was in interactivity and construction.
+The one activity that actually improved student understanding was not the viewing of a visualization, but instead its construction.
+
+If students, instead of viewing a visualization, construct their own, then they
+learn.  Unfortunately, visualizations are generally desired when a system is
+sufficiently complex to be difficult to understand.  If a system is difficult
+to understand, then it is difficult to program, and, even worse, graphics
+programming is perceived as one of the trickiest programming tasks.  This means
+that when we ask students to build a visualization, we are asking them to take
+a system they do not understand, build it correctly, and then engage in an
+extremely delicate programming task with their (hopefully correct, but likely
+buggy) system.
+
+For educational visualization to be effective, students must write their own
+visualizations.  Therefore, whenever possible, we design our system to be as
+simple, learnable, and robust as possible.  This will allow more students to
+succeed in the creation of visualizations by attemping to make the graphics
+aspect of the programming task as simple as possible.
 
@@
 Computational processes are often both sophisticated and beautiful, but their
-beauty is often lost in the nitty-gritty of simply getting a given algorithm to
-work.  Using a visualization system could, hopefully, remind students of the
+beauty can be lost in the nitty-gritty of simply getting a given algorithm to
+work.  Using a visualization system could, potentially, remind students of the
 beauty of the structures they are working with and also serve as an effective
 advertisement to non-majors of the intricacy and interestingness available in
-the CS major.
+the computer science and mathematics majors.
+
+To acheive this goal of beauty, we will pay attention to the aesthetic aspects
+of our visualization as well as maintaining correctness.  In particular, we
+should make sure that transitions are smooth and not jarring, and that our
+layouts are as aesthetically pleasing as possible.  This requirement is
+difficult to quantify, but, cetus paribus, a beautiful visualization is always
+better than an ugly one.
 
@@
 however, we found that every domain we considered had a naturally occuring root
-and ordering, and that the imposition of an arbitrary root and ordering was
-unnecessary.  From now on, when we say tree, we will mean a rooted ordered tree.
+and ordering, and that the imposition of an arbitrary root or an arbitrary
+ordering was unnecessary.  From here forward, when we say tree, we will mean a
+rooted ordered tree.
 
+\section{Constraints of Tree Layout}
+\label{sec:constraints}
 To lay out a tree in the plane, we must map every vertex $v \in V$ to a
@@
 Our last requirement is that if we have two nodes $w$ and $z$ with equal depth and children $U$ and $V$, and $w_x < z_x$, then $u_x < v_x \forall u\in U, v\in V$.
-This last requirement allows us to avoid crossing edges.
+This will avoid crossing edges, which is a good thing because trees are planar graphs, and so drawing them with crossing edges is totally unnecessary.
 
@@
 
-Paper from 1996 and Tamassia and the force-directed placement algorithm\cite{fdp}.
+Tree animation has a rich body of work associated with it.  In particular, much
+past work has focused on exactly what tree operations may be supported
+efficiently\cite{Moen,tamassia} as well as on the force-directed placement
+algorithm for tree layout\cite{fdp, Cohen}.  The entirety of this work has been
+largely spent on designing systems in which trees can be laid out and animated
+in an efficient manner.  Our work differs from theirs due to our focus on
+ease-of-use, and that focus comes from the fact that the speed of a ``current''
+computer has grown dramatically since the original papers were written.
 
@@
 library easy to use, even if it increases the polynomial runtime of some of our
-methods.  This is safe to do because we have restricted ourself to those graphs
-which can be drawn, in their entirety, on a computer screen, and have therefore
-restricted ourself to the (admittedly ill-defined) set of graphs which are
-``small enough'' to still be drawn quickly using algorithms which may be of
-order $\omega(n)$, as long as these algorithms are not super-polynomial.
+methods.  This is safe to do because we have restricted ourselves to those
+graphs which can be drawn in their entirety on a computer screen.  We have
+therefore restricted ourselves to the (admittedly ill-defined) set of graphs
+which are ``small enough'' to still be drawn quickly using algorithms which may
+be of order $\omega(n)$, as long as these algorithms are not super-polynomial.
 
 Therefore, instead of restricting ourself to a particular subset of allowed
-tree operations, we allow arbitrary transformations of the tree, and then we
+tree operations, we allow arbitrary transformations of the tree (arbitrary subtree creation, deletion, renaming, and movement), and then we
 animate our old tree transforming into the new tree.
@@
 
-1) Fix the root node to the middle top position. 
-2) For each node of a given level treat every 
-link as a spring and every node as a magnet 
-and calculate the resultant forces. 
-3) Simulate the change in position due to 
-these forces and update all node positions 
-by a small amount. 
-4) Display the system to the user, and repeat
+Our tree algorithm closely follows the force-directed layout algorithm of Fruchterman and Reingold\cite{fdp}, with some modifications particular to the layout of trees.
+In force-directed layout, vertices are treated as magnets which mutually repel each other and edges are treated are treated as springs which pull the adjacent vertices towards each other.
+The system is then allowed to evolve over time until a minimal energy state has been reached, resulting in a graph layout which is generally pleasing to the eye\cite{Cohen}.
+
+Our algorithm differs from the traditional method in two ways: we specify an initial layout, and we only allow movement in certain directions.
+Our initial layout is begins with a breadth-first search of the tree, beginning at the root, in which we search the children of a vertex in their natural order.
+From this, we construct our initial layout by putting all vertices of equal depth at the same y-coordinate, evenly spaced out horizontally.
+The horizontal order is determined by the order in which the nodes occur in the breadth-first search.
+Using this algorithm, we create an initial layout which obeys the constraints detailed in Section~\ref{sec:constraints}.
+
+Starting with this initial layout, we use force-directed layout, but we place
+constraints on how the vertices can move in response to the forces applied.
+Most importantly, we only allow vertices to left left-right, and never
+up-down.  The initial vertical position of a vertex is also its final
+position.  Similarly, a vertex may never swap rank with another vertex on the
+same level --- no matter how strong the forces involved, the initial ordering
+of a row must also be the final ordering of a row.  Metaphorically, one can
+think of treating the vertices as train cars on a horizontal rail.  The cars
+can only move back and forth in one direction, and they may never move through
+each other.
+
+In a large graph, this layout may take some time to settle down (on the order
+of seconds).  We have seen positive effects in displaying the layout process to
+the user.  In particular, multiple users have commented how, by showing the
+layout process settling the tree into the final layout, the graph seems more
+``alive''.
 
@@
 
-Either on every tree transformation, or when requested by the visualization
-programmer, we relayout the entire tree according to our layout algorithm, and
-then we animate the transition from the older, currently-laid-out tree, to the
-new tree.  Or animation proceeds in the following steps:
+Either on every tree transformation or when requested by the visualization
+programmer, we relayout (off-screen) the entire tree according to our layout
+algorithm and then we animate the transition from the old tree to the new
+tree.  Or animation procedure is extremely basic, but the simplicity allows for simplicity of use.
 
-1) Layout the new tree using the method above. 
-Do this layout off screen. 
-2) Fade out any of the nodes of the currently 
-displayed original tree that have been removed. 
-3) Compare the locations of the nodes in the 
-original tree to those in the new tree, then 
-animate the movement of the original nodes 
-to their new positions. 
-4) Fade in any nodes that only exist in the 
-new tree. 
-Multiple changes can be either combined into a 
-single animation step, or each step may be 
-animated individually.  This allows artistic flexibility 
-when creating a new visualization.
+Our animation procedure has six steps.  First, we lay out, off-screen, the new
+tree.  Second, we zoom out (if required) so that the viewable area is the size
+of the maximum of the new tree and old tree.  Third, we fade out any of the
+vertices and edges which were in the old tree, but have been removed in the new
+tree. Fourth, we animate the movement of all vertices which are present in both
+trees from their old positions to their new postions.  Fifth, we fade in any
+new nodes which are present only in the new tree.  Sixth, we zoom in (if
+required) so that the new tree fills the viewing area.
+
+Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default  This allows artistic flexibility when creating a new visualization, but allows visualization programmers to err on the side of more visualization rather than less.
 
@@
 
-Our library is written in Java, and users of the library need only use two of
+Our library\cite{ourlibrary} is written in Java, and users of the library need only use two of
 the classes it contains: the \java{Tree} class, and the \java{TreeDisplay}
@@
 code.  A \java{Tree} may be created at any time, and to display a tree, use the
-\java{setTree} method of the \java{TreeDisplay} class.
+\java{setTree} method of the \java{TreeDisplay} class.  Each \java{Tree} has
+methods to manipulate both the data displayed at the root of that particular
+subtree, as well as to manage the children of the root note of the subtree,
+which are, naturally, trees themselves.
 
-\subsection{A Sample Application}
+Each \java{Tree} has no ties to any algorithm.
+The user must program any manipulation algorithm themselves.
+If the \java{Tree} is currently being displayed, then any tree operation will be animated by default.
 
-\section{Example Animations}
-\subsection{Random Tree Growth}
-\subsection{Self-Balancing Binary Search Tree}
-\subsection{Tree Automaton}
+\subsection{Example: Tree Automaton}
+
+Our first example usage of the library is the animation of a tree automaton.
+Tree automatons are a long-studied branch of computational theory, and our tree automaton is adapted from the first chapter of {\it Tree Automata Techniques and Applications}\cite{tata2007}.
+
+The tree automaton we visualize accepts all trees which represent logical tautologies made out of the logical operations ``and'', ``or'', and ``not'', and the literals 0 and 1.
+Our rewrite rules for the tree are shown in Figure~\ref{fig:rules}, and the set of accepted trees is a tree with a single vertex named ``true''.
+
+\begin{figure}
+  \begin{tabular}{c c c}
+     1 $\to$ true            & (and true false) $\to$ false & (or true false) $\to$ true \\
+     0 $\to$ false           & (and true true) $\to$ true & (or true true) $\to$ true \\
+     (not true) $\to$ false & (and false false) $\to$ false & (or false false) $\to$ false \\
+     (not false) $\to$ true & (and false true) $\to$ false & (or false true) $\to$ true 
+    \end{tabular}
+\caption{The rewrite rules for a tree automaton which recognizes tautologies.
+We use s-expressions to denote parent-child relations.
+For example, the logical expression $1 \land 0$ would then be denoted (and 1 0), meaning that ``and'' is the parent vertex and 1 and 0 the child vertices.}
+\label{fig:rules}
+\end{figure}
+
+The progressive execution of the tree automaton consists of recognizing some
+subtree which conforms to the left hand side of a rule and then replacing that
+subtree with the right side of the rule.  Our visualization library alowed us
+to implement a tree automation and its attendant visualization using
+approximately 200 lines of Java code.
+The size is mentioned because a project of that size is small enough to be appropriate for a class assignment, which is a positive sign that we are meeting the requirements of Section~\ref{educ}.
+
+Our visualization of this automaton analyzing the logical expression \\
+~\\
+($\lnot$ (($\lnot$ ((1 $\lor$ 1) $\lor$ ($\lnot$ 1))) $\land$ ((($\lnot$ 1) $\land$ (0 $\land$ 0)) $\lor$ ((1 $\lor$ 1) $\lor$ (1 $\lor$ 0))))) \\
+$\land$ ($\lnot$ (((($\lnot$ 1) $\lor$ (0 $\land$ 1)) $\land$ ((0 $\lor$ 0) $\land$ (1 $\lor$ 1))) $\land$ (((1 $\land$ 0) $\lor$ (0 $\land$ 1)) $\lor$ ((1 $\land$ 1) $\lor$ (0 $\lor$ 0)))))\\
+~\\
+may be seen in flipbook form in Figure~\ref{fig:auto}.
+Interesting things of note include a few frames of the animation where a vertex is in the process of ``fading out'', as well as the successive zooming as the tree gets smaller and smaller.
+Note that our final frame indicates that the tree automaton has accepted the input, and that therefore the logical expression was a tautology.
+
 \begin{figure}
@@
 \end{center}
-\caption{Execution of a tree automaton which recognizes tautologies}
+\caption{A flipbook-style visualization of the execution of a tree automaton which recognizes tautologies generated by our library.}
+\label{fig:auto}
 \end{figure}
 
-\subsection{Lisp}
+Our visualization of a tree automaton was, in our own estimation,
+straightforward to create, and it does provide a visualization of the process
+of an automaton attempting to recognize a tree.
+The animation proceeds in a smooth manner on a laptop purchased in 2008, which indicates that, for a problem of this size, our library is practically useful as well as theoretically fast enough.
+
+\subsection{Example: Lisp}
+
+We also created a visualization of a lisp machine executing a very basic program.
+Lisp programs are represented by s-expressions, and these s-expressions directly correspond to a pre-order traversal of a tree.
+Thus, we represent the lisp program which calculates one of the roots of $x^2+x-6$ as shown in Figure~\ref{lispprog}.
+\begin{figure}
+\begin{verbatim}((fun (a b c) 
+    (/  (+  (* -1 b) 
+            (^ (- (^ b 2) (* 4 (* a c))) .5)) 
+        (* 2 a)))  
+1 1 -6)
+\end{verbatim}
+\caption{A Lisp program which calculates one of the roots of $x^2+x-6$ using the quadratic formula.}
+\label{lispprog}
+\end{figure}
+
+There exist multiple virtual machines which model the execution of lisp code in equivalent manners, here we used one based on string substitution.
+The interested reader is referred to {\it Abstract computing Machines} by Kluge\cite{acm}.
+This particular visualization took more code than the tree automaton, but it includes the implementation of a turing-complete lisp compiler and virtual machine, and so may be forgiven for the extra length.
+The total length of all the code required was one the order of 450 lines of Java, and so is suitable for a long project or a perhaps a group project in a programming languages course.
+
+The results of our visualization may be found in the animation flipbook shown
+in Figure~\ref{lispanim}. In the flipbook, one can see many examples of
+vertices caught in the middle of an animation.  There are multiple examples of
+vertices fading out or caught in the middle of moving around the screen.
 
@@
 \caption{The complete execution of a lisp program calculating one of the roots of $x^2+x-6$.}
+\label{lispanim}
 \end{figure}
+
+As well as the execution of the evaluation of the quadratic formula, we have also produced visualizations of lisp code calculating the factorial function and the Fibonacci sequence.
+These examples, while interesting, are mostly interesting as examples of the $\lambda$-calculus and teaching aids in a programming languages course.
+The animation proceeds much like the the flipbook shown, but the trees involved are much larger and more inctricate.
+
+\subsection{Other example visualizations}
+
+Also included with our library are visualizations of self-adjusting binary search trees (AVL Trees) and of the growth of a random tree.
+The random tree is particularly interesting, because it was requested by an audience member during a presentation and was then implemented in the next 30 minutes.
+It is hoped that this short distance from request to implementation can serve as an example of the power of both our library and especially our design approach.
 
@@
 layout library that, by doing much more work behind the scenes, makes the
-creation and usage of visualizations available to a much larger audience.  We
-have created such a library and provided several sample visualizations.  We
-hope that any and all of our design ideas, the resulting library, and the
-provided visualizations, might be useful to others who are attempting to
-demonstrate, learn, understand, or admire the behavior of trees which evolve
+creation of visualizations available to a much larger audience.  We have
+created such a library and provided several sample visualizations.  We hope
+that any and all of our design ideas, the resulting library\cite{ourlibrary},
+and the provided visualizations, might be useful to others who are attempting
+to demonstrate, learn, understand, or admire the behavior of trees which evolve
 over time.
 
-\bibliographystyle{plain}
+\bibliographystyle{abbrv}
 \bibliography{bibliography.bib}

### 2010-05-14 paper/paper.tex
Possibly the one I am submitting in an hour.

@@
 to be displayed on a screen, computers are now fast enough to eliminate many
-worries about efficiency.  This allows us to make a library which emphasizes
-simplicity of use over efficiency of execution.  We demonstrate our library
-using several visualizations.\end{abstract}
+worries about efficiency.  This observation allows us to make a library which
+emphasizes simplicity of use over efficiency of execution.  We show how to use  our
+library using several visualizations.\end{abstract}
 
@@
 build a tree visualization system which might be useful in many different
-contexts.  In this paper we articulate the design principles of a highly usable
+contexts.  We will articulate the design principles of a highly usable
 tree visualization system, present one such system that we have built in Java
-and made freely available, and then demonstrate our system's utility with
+and made freely available, and then show our system's utility with
 several sample visualizations.
@@
 
-Computing and math eductors have a long history of building visualization
-systems for demonstration, on the grounds that if students can view these
+Computing and math educators have a long history of building visualization
+systems for demonstration.  The intuition in many case was that if students can view these
 systems in action, then the students will understand these system.
@@
 The one bright spot in this was in interactivity and construction.
-The one activity that actually improved student understanding was not the viewing of a visualization, but instead its construction.
+The one activity that improved student understanding was not the viewing of a visualization, but instead its construction.
 
@@
 
-For educational visualization to be effective, students must write their own
+For effective educational visualization, students must write their own
 visualizations.  Therefore, whenever possible, we design our system to be as
 simple, learnable, and robust as possible.  This will allow more students to
-succeed in the creation of visualizations by attemping to make the graphics
+succeed in the creation of visualizations by attempting to make the graphics
 aspect of the programming task as simple as possible.
@@
 Computational processes are often both sophisticated and beautiful, but their
-beauty can be lost in the nitty-gritty of simply getting a given algorithm to
+beauty can be lost in the nitty-gritty of simply making a given algorithm
 work.  Using a visualization system could, potentially, remind students of the
@@
 
-To acheive this goal of beauty, we will pay attention to the aesthetic aspects
+To achieve this goal of beauty, we will pay attention to the aesthetic aspects
 of our visualization as well as maintaining correctness.  In particular, we
-should make sure that transitions are smooth and not jarring, and that our
+should ensure that transitions are smooth and not jarring, and that our
 layouts are as aesthetically pleasing as possible.  This requirement is
@@
 length of the shortest path between $v$ and $r$. Every edge connects two
-vertices $(u,v)$, and we will assume, again wlog, that the depth of $u$ is one
+vertices $(u,v)$, and we will assume, again without loss of generality, that the depth of $u$ is one
 less than the depth of $v$.  In that case, we call $u$ the \newword{parent} of
@@
 selecting a root and imposing a partial-ordering on the vertices.  In practice,
-however, we found that every domain we considered had a naturally occuring root
+however, we found that every domain we considered had a naturally occurring root
 and ordering, and that the imposition of an arbitrary root or an arbitrary
@@
 position.  Similarly, a vertex may never swap rank with another vertex on the
-same level --- no matter how strong the forces involved, the initial ordering
+same level. No matter how strong the forces involved, the initial ordering
 of a row must also be the final ordering of a row.  Metaphorically, one can
@@
 the user.  In particular, multiple users have commented how, by showing the
-layout process settling the tree into the final layout, the graph seems more
-``alive''.
+layout process settling the tree into the final layout, the graph seemed more
+``alive'' on the screen.
 
@@
 Either on every tree transformation or when requested by the visualization
-programmer, we relayout (off-screen) the entire tree according to our layout
+programmer, we re-layout (off-screen) the entire tree according to our layout
 algorithm and then we animate the transition from the old tree to the new
@@
 tree. Fourth, we animate the movement of all vertices which are present in both
-trees from their old positions to their new postions.  Fifth, we fade in any
+trees from their old positions to their new positions.  Fifth, we fade in any
 new nodes which are present only in the new tree.  Sixth, we zoom in (if
@@
 
-Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default  This allows artistic flexibility when creating a new visualization, but allows visualization programmers to err on the side of more visualization rather than less.
+Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default  This choice allows artistic flexibility when creating a new visualization, but encourages visualization programmers to err on the side of more visualization rather than less.
 
@@
 by another programmer.  Therefore, our two classes have very few methods, and
-those few methods are very simple to use.  Prior efforts focused a lot of work
+those few methods are very simple to use.  Prior efforts focused 
 on making specific animations efficient, but we have instead focused our
@@
 methods to manipulate both the data displayed at the root of that particular
-subtree, as well as to manage the children of the root note of the subtree,
-which are, naturally, trees themselves.
+subtree, as well as methods to manage the children of the subtree.
+Those children are, naturally, trees themselves.
 
@@
 The user must program any manipulation algorithm themselves.
-If the \java{Tree} is currently being displayed, then any tree operation will be animated by default.
+If the \java{Tree} is being actively displayed, then any tree operation will be animated by default.
 
@@
 The tree automaton we visualize accepts all trees which represent logical tautologies made out of the logical operations ``and'', ``or'', and ``not'', and the literals 0 and 1.
-Our rewrite rules for the tree are shown in Figure~\ref{fig:rules}, and the set of accepted trees is a tree with a single vertex named ``true''.
+Our rewrite rules for the tree may be found in Figure~\ref{fig:rules}, and the set of accepted trees is a tree with a single vertex named ``true'' and no edges.
 
@@
 subtree which conforms to the left hand side of a rule and then replacing that
-subtree with the right side of the rule.  Our visualization library alowed us
+subtree with the right side of the rule.  Our visualization library allowed us
 to implement a tree automation and its attendant visualization using
@@
 Interesting things of note include a few frames of the animation where a vertex is in the process of ``fading out'', as well as the successive zooming as the tree gets smaller and smaller.
-Note that our final frame indicates that the tree automaton has accepted the input, and that therefore the logical expression was a tautology.
+The final frame of the flipbook indicates that the tree automaton has accepted the input, and that therefore the logical expression was a tautology.
 
@@
 The interested reader is referred to {\it Abstract computing Machines} by Kluge\cite{acm}.
-This particular visualization took more code than the tree automaton, but it includes the implementation of a turing-complete lisp compiler and virtual machine, and so may be forgiven for the extra length.
+This particular visualization took more code than the tree automaton, but it includes the implementation of a Turing-complete lisp compiler and virtual machine, and so may be forgiven for the extra length.
 The total length of all the code required was one the order of 450 lines of Java, and so is suitable for a long project or a perhaps a group project in a programming languages course.
@@
 in Figure~\ref{lispanim}. In the flipbook, one can see many examples of
-vertices caught in the middle of an animation.  There are multiple examples of
-vertices fading out or caught in the middle of moving around the screen.
+vertices caught in the middle of an animation, either fading out or in the process of moving about the screen.
 
@@
 
-As well as the execution of the evaluation of the quadratic formula, we have also produced visualizations of lisp code calculating the factorial function and the Fibonacci sequence.
-These examples, while interesting, are mostly interesting as examples of the $\lambda$-calculus and teaching aids in a programming languages course.
-The animation proceeds much like the the flipbook shown, but the trees involved are much larger and more inctricate.
+As well as the execution of the evaluation of the quadratic formula, we have
+also produced visualizations of lisp code calculating the factorial function
+and the Fibonacci sequence.  These examples, while interesting, are mostly
+interesting as examples of the $\lambda$-calculus and teaching aids in a
+programming languages course.  The animation proceeds much like the flipbook
+shown, but the trees involved are much larger and more intricate.
 
-\subsection{Other example visualizations}
+\subsection{Other Example Visualizations}
 
-Also included with our library are visualizations of self-adjusting binary search trees (AVL Trees) and of the growth of a random tree.
-The random tree is particularly interesting, because it was requested by an audience member during a presentation and was then implemented in the next 30 minutes.
-It is hoped that this short distance from request to implementation can serve as an example of the power of both our library and especially our design approach.
+Also included with our library are visualizations of self-adjusting binary
+search trees (AVL Trees) and of the growth of a random tree.  The random tree
+is particularly interesting, because it was requested by an audience member
+during a presentation and was then implemented in the next half hour.  We hope
+that this short distance from request to implementation can serve as an example
+of the power of both our library and especially our design approach.
 

### 2010-05-14 paper/paper.tex
Submitted.

@@
 computer algorithms.  It would be useful to be able to easily draw trees and
-animate them changing over time.  We begin by articulating a design for tree
+animate them changing over time.  We articulate a design for tree
 visualization and animation that is more general than previous methodologies.
@@
 to be displayed on a screen, computers are now fast enough to eliminate many
-worries about efficiency.  This observation allows us to make a library which
+concerns about efficiency.  This observation allows us to make a library which
 emphasizes simplicity of use over efficiency of execution.  We show how to use  our
@@
 computer science.  They occur in many disparate contexts, and while the uses of
-these trees may be different, there exist commonalities which we can use to
+these trees may be different, there exist commonalities which we exploit to
 build a tree visualization system which might be useful in many different
@@
 we discuss the particulars of our visualization, we must explain our reasons
-for wanting such a system in the first place, and the expected audience.  We
-want to visualize dynamic trees for demonstration, education, and beauty, and
-we would like our visualizations to be of benefit to computer science
-undergraduates.
+for wanting such a system in the first place, as well as identify the expected
+audience.  We want to visualize dynamic trees for demonstration, education, and
+beauty, and we would like our system to be of benefit to computer
+science undergraduates.
 
@@
 
-One reason for visualization is for demonstration.  Trees (and other systems)
+One reason for visualization is for demonstration.  Trees and other systems
 are visualized in an effort to show how these systems work.  Because this is
@@
 Computing and math educators have a long history of building visualization
-systems for demonstration.  The intuition in many case was that if students can view these
-systems in action, then the students will understand these system.
+systems for demonstration.  The intuition was that if students could view these
+systems in action, then the students would understand these systems.
 Unfortunately for those goals, Hundhausen, Douglas, and Stasko found that, in practice, almost all
-visualizations are useless\cite{hds}.  A good visualization seems intuitively useful, but
+visualizations are useless\cite{hds}.  A good visualization seems intuitively helpful, but
 it turns out that if a student does not understand the system being visualized, they will not understand the visualization either.
 The one bright spot in this was in interactivity and construction.
-The one activity that improved student understanding was not the viewing of a visualization, but instead its construction.
+The activity that improved student understanding was not the viewing of a visualization, but instead its construction and interactive use.
 
@@
 to understand, then it is difficult to program, and, even worse, graphics
-programming is perceived as one of the trickiest programming tasks.  This means
+programming can be one of the trickiest programming tasks.  This means
 that when we ask students to build a visualization, we are asking them to take
@@
 simple, learnable, and robust as possible.  This will allow more students to
-succeed in the creation of visualizations by attempting to make the graphics
+succeed in the creation of visualizations by making the graphics
 aspect of the programming task as simple as possible.
@@
 
-To achieve this goal of beauty, we will pay attention to the aesthetic aspects
-of our visualization as well as maintaining correctness.  In particular, we
-should ensure that transitions are smooth and not jarring, and that our
-layouts are as aesthetically pleasing as possible.  This requirement is
-difficult to quantify, but, cetus paribus, a beautiful visualization is always
-better than an ugly one.
+This requirement is difficult to quantify, but, cetus paribus, a beautiful
+visualization is always better than an ugly one.  To achieve this goal of
+beauty, we will pay attention to the aesthetic aspects of our visualization as
+well as maintaining correctness.  In particular, we should ensure that
+transitions are smooth and not jarring, and that our layouts are as
+aesthetically pleasing as possible.
 
@@
 length of the shortest path between $v$ and $r$. Every edge connects two
-vertices $(u,v)$, and we will assume, again without loss of generality, that the depth of $u$ is one
+vertices $(u,v)$, and we will assume, without loss of generality, that the depth of $u$ is one
 less than the depth of $v$.  In that case, we call $u$ the \newword{parent} of
@@
 
-Without loss of generalization, we restrict our visualization system to rooted,
+We restrict our visualization system to rooted,
 ordered trees.  Any tree can be converted into a rooted, ordered tree by
@@
 ordering was unnecessary.  From here forward, when we say tree, we will mean a
-rooted ordered tree.
+rooted, ordered tree.
 
@@
 Our last requirement is that if we have two nodes $w$ and $z$ with equal depth and children $U$ and $V$, and $w_x < z_x$, then $u_x < v_x \forall u\in U, v\in V$.
-This will avoid crossing edges, which is a good thing because trees are planar graphs, and so drawing them with crossing edges is totally unnecessary.
+This will avoid crossed edges, which is a good thing because trees are planar graphs, and so drawing them with crossing edges is totally unnecessary.
 
@@
 in an efficient manner.  Our work differs from theirs due to our focus on
-ease-of-use, and that focus comes from the fact that the speed of a ``current''
-computer has grown dramatically since the original papers were written.
+ease-of-use, and that focus comes from the fact that the speed of commonly
+available computers has grown dramatically since the original papers were
+written.
 
@@
 programmer, we re-layout (off-screen) the entire tree according to our layout
-algorithm and then we animate the transition from the old tree to the new
-tree.  Or animation procedure is extremely basic, but the simplicity allows for simplicity of use.
+algorithm and then we animate the transition from the old tree to the new tree.
+Or animation procedure is extremely simple, but that simplicity allows for
+simplicity of use.
 
@@
 tree.  Second, we zoom out (if required) so that the viewable area is the size
-of the maximum of the new tree and old tree.  Third, we fade out any of the
+of the maximum required for the new tree and old tree.  Third, we fade out any of the
 vertices and edges which were in the old tree, but have been removed in the new
@@
 
-Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default  This choice allows artistic flexibility when creating a new visualization, but encourages visualization programmers to err on the side of more visualization rather than less.
+Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default.
+This choice allows artistic flexibility when creating a new visualization, but encourages visualization programmers to err on the side of more visualization rather than less.
 
